---
title: "Epilepsy MapMyCells Annotation"
output: html_notebook
---

This script prepares the epilepsy Seurat object for MapMyCells annotation. Since no annotations are currently present in the object, we will process all cells and let MapMyCells identify neurons.

## 1. Load Libraries

```{r}
# Load Packages
library(Seurat)
library(rliger)
library(scCustomize)
library(qs)
library(anndata)
library(reticulate)
library(dplyr)
library(ggplot2)

# Install Python anndata module
reticulate::py_install("anndata")
```

## 2. Load and Inspect Epilepsy Object

```{r}
# Load the epilepsy Seurat object
obj_path <- "D:/Youssef Sibih/Seurat_Objects/R1_123125/epilepsy_4samples_combined_sct.rds"
obj <- readRDS(obj_path)

# Basic object information
cat("Object loaded successfully!\n")
cat("Total cells:", ncol(obj), "\n")
cat("Total genes:", nrow(obj), "\n\n")

# Check Seurat version
cat("Seurat version:", as.character(packageVersion("Seurat")), "\n")
cat("Object class:", paste(class(obj), collapse = ", "), "\n\n")

# List available assays
cat("Available assays:", names(obj@assays), "\n")
cat("Default assay:", DefaultAssay(obj), "\n\n")

# Check active identities
cat("Active identities (Idents):\n")
print(table(Idents(obj)))
cat("\nNote: No annotations present - processing all cells\n\n")

# List metadata columns
cat("Available metadata columns:\n")
print(colnames(obj@meta.data))
cat("\n")

# Check for seurat_clusters or other grouping variables
if ("seurat_clusters" %in% colnames(obj@meta.data)) {
  cat("seurat_clusters found:\n")
  print(table(obj@meta.data$seurat_clusters))
  cat("\n")
} else {
  cat("seurat_clusters not found in metadata\n\n")
}

# Check if UMAP or other reductions are available
cat("Available reductions:", names(obj@reductions), "\n\n")

# Display basic visualization if UMAP exists
if ("umap" %in% names(obj@reductions)) {
  DimPlot(obj, reduction = "umap", label = TRUE) + 
    ggtitle("Epilepsy Object - All Cells (No Annotations)")
} else if (length(obj@reductions) > 0) {
  # Use first available reduction
  reduction_name <- names(obj@reductions)[1]
  DimPlot(obj, reduction = reduction_name, label = TRUE) + 
    ggtitle(paste("Epilepsy Object - All Cells (", reduction_name, ")", sep = ""))
}
```

## 3. Convert All Cells to h5ad Format for MapMyCells

Since there are no annotations in the object, we will process all cells. MapMyCells will identify and annotate neurons from all cells.

```{r}
# Set default assay to RNA (required for MapMyCells)
DefaultAssay(obj) <- "RNA"

# Check if object uses v5 layers (needs JoinLayers) or v3 (no layers)
# Try to join layers - will work if v5, will error if v3
tryCatch({
  obj_joined <- JoinLayers(obj, assay = "RNA")
  cat("Successfully joined layers (v5 Seurat object)\n")
  use_joined <- TRUE
}, error = function(e) {
  cat("JoinLayers not needed or failed (likely v3 Seurat object):", e$message, "\n")
  cat("Using object as-is\n")
  use_joined <- FALSE
})

# Use joined object if available, otherwise use original
if (exists("obj_joined") && use_joined) {
  obj_for_conversion <- obj_joined
} else {
  obj_for_conversion <- obj
}

# Check object size to determine if splitting is needed
cell_count <- ncol(obj_for_conversion)
cat("\nTotal cells to convert:", cell_count, "\n")

# Define output directory (MapMyCells_Input subdirectory in R1_123125)
output_dir <- file.path(dirname(obj_path), "MapMyCells_Input")
# Create directory if it doesn't exist
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
  cat("Created output directory:", output_dir, "\n")
}
cat("Output directory:", output_dir, "\n\n")

# Determine if splitting is needed
# MapMyCells can handle large files, but if >100k cells, consider splitting
SPLIT_THRESHOLD <- 100000  # Adjust based on your system's capabilities

if (cell_count > SPLIT_THRESHOLD && "seurat_clusters" %in% colnames(obj_for_conversion@meta.data)) {
  cat("Object is large (", cell_count, " cells). Splitting into groups...\n\n")
  
  # Ensure seurat_clusters is character
  obj_for_conversion$seurat_clusters <- as.character(obj_for_conversion$seurat_clusters)
  
  # Get unique clusters
  unique_clusters <- unique(obj_for_conversion$seurat_clusters)
  cat("Found", length(unique_clusters), "clusters\n")
  
  # Split into two groups (adjust logic as needed)
  # Simple approach: split clusters roughly in half
  mid_point <- ceiling(length(unique_clusters) / 2)
  group1_clusters <- unique_clusters[1:mid_point]
  group2_clusters <- unique_clusters[(mid_point + 1):length(unique_clusters)]
  
  cat("Group 1 clusters:", paste(group1_clusters, collapse = ", "), "\n")
  cat("Group 2 clusters:", paste(group2_clusters, collapse = ", "), "\n\n")
  
  # Subset Group 1
  obj_group1 <- subset(obj_for_conversion, seurat_clusters %in% group1_clusters)
  cat("Group 1 cells:", ncol(obj_group1), "\n")
  
  # Subset Group 2
  obj_group2 <- subset(obj_for_conversion, seurat_clusters %in% group2_clusters)
  cat("Group 2 cells:", ncol(obj_group2), "\n\n")
  
  # Convert Group 1 to h5ad
  cat("Converting Group 1 to h5ad format...\n")
  as.anndata(
    x = obj_group1, 
    file_path = output_dir, 
    file_name = "epilepsy_anndata_group1.h5ad"
  )
  cat("Saved: epilepsy_anndata_group1.h5ad\n\n")
  
  # Convert Group 2 to h5ad
  cat("Converting Group 2 to h5ad format...\n")
  as.anndata(
    x = obj_group2, 
    file_path = output_dir, 
    file_name = "epilepsy_anndata_group2.h5ad"
  )
  cat("Saved: epilepsy_anndata_group2.h5ad\n\n")
  
  cat("Conversion complete! Two files created for MapMyCells processing.\n")
  
} else {
  # Single file conversion
  cat("Converting all cells to single h5ad file...\n")
  
  as.anndata(
    x = obj_for_conversion, 
    file_path = output_dir, 
    file_name = "epilepsy_4samples_combined_anndata.h5ad"
  )
  
  cat("Saved: epilepsy_4samples_combined_anndata.h5ad\n")
  cat("Conversion complete! File ready for MapMyCells processing.\n")
}
```

## 4. Summary

```{r}
cat("=== Conversion Summary ===\n")
cat("Input object:", basename(obj_path), "\n")
cat("Total cells processed:", ncol(obj), "\n")
cat("Output directory:", output_dir, "\n")

if (cell_count > SPLIT_THRESHOLD && "seurat_clusters" %in% colnames(obj_for_conversion@meta.data)) {
  cat("Files created:\n")
  cat("  - epilepsy_anndata_group1.h5ad\n")
  cat("  - epilepsy_anndata_group2.h5ad\n")
} else {
  cat("File created:\n")
  cat("  - epilepsy_4samples_combined_anndata.h5ad\n")
}

cat("\nNext steps:\n")
cat("1. Open the h5ad file(s) in MapMyCells GUI\n")
cat("2. Use MTG atlas to annotate neurons\n")
cat("3. Export annotations as CSV\n")
cat("4. Use the next script to import annotations back to Seurat object\n")
```

