---
title: "Epilepsy MapMyCells Plots and Analysis"
output: html_notebook
---

This script loads MapMyCells annotations, integrates them into the epilepsy Seurat object, generates plots, and exports analysis data.

## 1. Load Libraries

```{r}
library(Seurat)
library(dplyr)
library(ggplot2)
library(ggpattern)
library(scales)
library(ggpubr)
library(RColorBrewer)
library(patchwork)
```

## 2. Load Seurat Object and MapMyCells Annotations

```{r}
# Load the epilepsy Seurat object
obj_path <- "D:/Youssef Sibih/Seurat_Objects/R1_123125/epilepsy_4samples_combined_sct.rds"
obj <- readRDS(obj_path)

cat("Object loaded successfully!\n")
cat("Total cells:", ncol(obj), "\n\n")

# Define mapping files for the two groups
mapping_files <- list(
  group1 = "D:/Youssef Sibih/Seurat_Objects/R1_123125/MapMyCells_Input/epilepsy_anndata_group1_10xHumanMTGSEA/epilepsy_anndata_group1_10xHumanMTGSEA-AD(CCN20230505)_SEA-ADHierarchicalMapping_UTC_1767498963499.csv",
  group2 = "D:/Youssef Sibih/Seurat_Objects/R1_123125/MapMyCells_Input/epilepsy_anndata_group2_10xHumanMTGSEA/epilepsy_anndata_group2_10xHumanMTGSEA-AD(CCN20230505)_SEA-ADHierarchicalMapping_UTC_1767501091103.csv"
)

# Load and combine both annotations
all_mappings <- lapply(mapping_files, function(file) {
  mapping <- read.csv(file, comment.char = "#")
  rownames(mapping) <- mapping$cell_id  # Ensure rownames are set
  return(mapping[, c("class_name", "subclass_name")])  # Keep only relevant columns
})

# Merge both group mappings
combined_mapping <- bind_rows(all_mappings)

# Ensure only matching cell IDs are used
common_cells <- intersect(rownames(obj@meta.data), rownames(combined_mapping))
cat("Found", length(common_cells), "cells with MapMyCells annotations\n")

# Subset combined mapping to include only cells in object
combined_mapping <- combined_mapping[common_cells, , drop = FALSE]

# Add the MapMyCells annotations to the original Seurat object
obj <- AddMetaData(obj, metadata = combined_mapping)

# Verify metadata integration
cat("MapMyCells annotations added to object\n")
cat("Unique class names:", paste(unique(obj@meta.data$class_name[!is.na(obj@meta.data$class_name)]), collapse = ", "), "\n\n")

# Visualize UMAP with new annotations (all cells)
DimPlot(obj, reduction = "umap", group.by = "class_name", label = TRUE) + 
  ggtitle("All Cells - MapMyCells Class")
```

## 3. Create Neuron Subset

```{r}
# Filter to neuronal cells only (Glutamatergic and GABAergic)
neuron_obj_cleaned <- subset(x = obj, subset = class_name %in% c("Neuronal: Glutamatergic", "Neuronal: GABAergic"))

cat("Neuron subset created\n")
cat("Total neurons:", ncol(neuron_obj_cleaned), "\n")
cat("Neuron subtypes:\n")
print(table(neuron_obj_cleaned@meta.data$subclass_name))
```

## 4. Define Color Schemes

```{r}
# Define the color mapping for each cell type
my_colors <- c(
  "Neuronal: Glutamatergic" = "orange",
  "Neuronal: GABAergic"     = "lightblue"
)

my_colors_2 <- c(
  "L2/3 IT"    = "#FEC89A",
  "L4 IT"      = "#FCD5CE",
  "L5 IT"      = "#FFDAB9",
  "L6 IT"      = "#F49595",
  "L6 IT Car3" = "#F1CEBE",
  "L5 ET"      = "#F3D17C",
  "L6 CT"      = "#FCEFB4",
  "L5/6 NP"    = "#F9ED85",
  "L6b"        = "#FFCAE9",
  "Chandelier" = "#A9E4DE",
  "Sst Chodl"  = "#83B8C6",
  "Lamp5"      = "#D0E444",    
  "Lamp5 Lhx6" = "#D0EDEF",
  "Pax6"       = "#E1D3F8",
  "Vip"        = "#B0B5ED", 
  "Sncg"       = "#D6DFE8",
  "Pvalb"      = "#8FC0E3",    
  "Sst"        = "#7FB3D5"    
)
```

## 5. Add Module Scores

```{r}
# Define excitatory and inhibitory gene lists
excitatory_list <- list(c("SLC17A6", "SLC17A7", "SLC17A8", "GRIA1", "GRIA2", "GRIA3", "GRIA4", 
                          "GRIN1", "GRIN2A", "GRIN2B", "GRIN2C", "GRIN2D", "GRIN3A", "GRIN3B", 
                          "GRM1", "GRM2", "GRM3", "GRM4", "GRM5", "GRM6", "GRM7", "GRM8", 
                          "TBR1", "FEZF2", "NEUROD1", "CTIP2", "SATB2", "CAMK2A", "CAMK2B", 
                          "NRGN", "SYT1", "PSD95", "DLG4", "SHANK1", "SHANK2", "SHANK3", 
                          "HOMER1", "HOMER2", "HOMER3", "VAMP1", "VAMP2"))

inhibitory_list <- list(c("GAD1", "GAD2", "SLC32A1", "SLC6A1", "GABRA1", "GABRA2", "GABRA3", 
                          "GABRA4", "GABRA5", "GABRA6", "GABRB1", "GABRB2", "GABRB3", 
                          "GABRG1", "GABRG2", "GABRG3", "GABBR1", "GABBR2", "DLX1", "DLX2", 
                          "DLX5", "DLX6", "ARX", "LHX6", "SST", "PVALB", "CALB1", "CALB2", 
                          "VIP", "CR", "NPY", "GPHN", "NKCC1", "SLC12A2", "KCC2", "SLC12A5"))

# Add module scores
neuron_obj_cleaned <- AddModuleScore(neuron_obj_cleaned, features = excitatory_list, 
                                     name = "Excitatory_Profile", nbin = 24)

neuron_obj_cleaned <- AddModuleScore(neuron_obj_cleaned, features = inhibitory_list, 
                                     name = "Inhibitory_Profile", nbin = 24)

cat("Module scores calculated\n")
```

## 6. Generate UMAP Plots

```{r}
# Create output directory for plots in the oneoverf_R1_10326 folder
plot_dir <- "D:/AbrahamDada/oneoverf/oneoverf_R1_10326/plots"
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}

# Plot 1: All cells - with labels
p1 <- DimPlot(obj, reduction = "umap", group.by = "subclass_name", 
              label = TRUE, repel = TRUE) + 
  NoAxes() +
  theme(legend.position = "right")

ggsave(filename = "epilepsy_allcells_labeled.pdf", 
       plot = p1, 
       path = plot_dir, 
       width = 9, height = 4, 
       device = "pdf", dpi = 600)

p1

# Plot 2: All cells - without labels
p2 <- DimPlot(obj, reduction = "umap", group.by = "subclass_name", 
              label = FALSE) + 
  NoAxes() +
  NoLegend()

ggsave(filename = "epilepsy_allcells_unlabeled.pdf", 
       plot = p2, 
       path = plot_dir, 
       width = 5.5, height = 3.6, 
       device = "pdf", dpi = 600)

p2

# Ensure subclass_name is a factor with sorted levels for legend (neuron subset)
neuron_obj_cleaned$subclass_name <- factor(neuron_obj_cleaned$subclass_name, 
                                           levels = sort(unique(neuron_obj_cleaned$subclass_name)))

# Plot 3: Neuron only - with labels
p3 <- DimPlot(neuron_obj_cleaned, reduction = "umap", group.by = "subclass_name", 
              cols = my_colors_2, label = TRUE, repel = TRUE) + 
  NoAxes() +
  theme(legend.position = "right")

ggsave(filename = "epilepsy_neurononly_labeled.pdf", 
       plot = p3, 
       path = plot_dir, 
       width = 9, height = 4, 
       device = "pdf", dpi = 600)

p3

# Plot 4: Neuron only - without labels
p4 <- DimPlot(neuron_obj_cleaned, reduction = "umap", group.by = "subclass_name", 
              cols = my_colors_2, label = FALSE) + 
  NoAxes() +
  NoLegend()

ggsave(filename = "epilepsy_neurononly_nolabel.pdf", 
       plot = p4, 
       path = plot_dir, 
       width = 5.5, height = 3.6, 
       device = "pdf", dpi = 600)

p4

# Check if one_over_f_status exists before plotting
if ("one_over_f_status" %in% colnames(neuron_obj_cleaned@meta.data)) {
  # Plot 5: Neuron only - colored by 1/f status with labels
  p5 <- DimPlot(neuron_obj_cleaned, reduction = "umap", group.by = "one_over_f_status", 
                label = TRUE, repel = TRUE) + 
    NoAxes() +
    theme(legend.position = "right")
  
  ggsave(filename = "epilepsy_neurononly_1overf_labeled.pdf", 
         plot = p5, 
         path = plot_dir, 
         width = 9, height = 4, 
         device = "pdf", dpi = 600)
  
  p5
  
  # Plot 6: Neuron only - colored by 1/f status without labels
  p6 <- DimPlot(neuron_obj_cleaned, reduction = "umap", group.by = "one_over_f_status", 
                label = FALSE) + 
    NoAxes() +
    NoLegend()
  
  ggsave(filename = "epilepsy_neurononly_1overf_nolabel.pdf", 
         plot = p6, 
         path = plot_dir, 
         width = 5.5, height = 3.6, 
         device = "pdf", dpi = 600)
  
  p6
} else {
  cat("Note: one_over_f_status not found in metadata. Skipping 1/f status plots.\n")
}
```

## 7. Proportion Bar Plot by 1/f Status

```{r}
# Check if one_over_f_status exists
if (!"one_over_f_status" %in% colnames(neuron_obj_cleaned@meta.data)) {
  warning("one_over_f_status not found in metadata. Skipping proportion plot.")
} else {
  # Create proportion table
  prop_df <- table(neuron_obj_cleaned$one_over_f_status,
                   neuron_obj_cleaned$subclass_name) |>
    as.data.frame() |>
    `colnames<-`(c("one_over_f_status", "type", "Freq")) |>
    group_by(one_over_f_status) |>
    mutate(prop = Freq / sum(Freq)) |>
    ungroup()
  
  # Set factor levels
  prop_df$type <- factor(prop_df$type, levels = rev(names(my_colors_2)))
  
  # Recode 1/f status labels
  prop_df <- prop_df |>
    mutate(one_over_f_status = recode(one_over_f_status,
                                      "low1overf"  = "Flat 1/f",
                                      "high1overf" = "Steep 1/f"),
           one_over_f_status = factor(one_over_f_status,
                                      levels = c("Flat 1/f", "Steep 1/f")))
  
  # Generate bar plot with pattern
  p4 <- ggplot(prop_df,
         aes(x      = prop,
             y      = type,
             fill   = type,
             pattern        = one_over_f_status,
             pattern_colour = one_over_f_status)) +
    
    geom_bar_pattern(
      stat     = "identity",
      position = position_dodge2(width = 0.9, reverse = TRUE),
      width    = 0.85,
      colour   = "black",
      pattern          = "stripe",
      pattern_angle    = 45,
      pattern_density  = 0.4,
      pattern_spacing  = 0.03,
      pattern_size     = 0.7,
      pattern_fill     = NA
    ) +
    
    geom_text(aes(label = label_number(accuracy = 0.001)(prop)),
              position = position_dodge2(width = 0.9, reverse = TRUE),
              hjust    = -0.15,
              vjust    = 0.5,
              size     = 3) +
    
    scale_fill_manual(values = my_colors_2, guide = "none") +
    scale_pattern_manual(values = c("Flat 1/f" = "stripe", "Steep 1/f" = "none"),
                         guide = "none") +
    scale_pattern_colour_manual(values = c("Flat 1/f" = "black",
                                           "Steep 1/f" = NA),
                                 guide = "none") +
    
    labs(x = "Proportion", y = NULL, title = "Neuron Subtype Proportions by 1/f Status") +
    theme_classic(base_size = 11) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none")
  
  ggsave(filename = "epilepsy_proportions_1overf.pdf", 
         plot = p4, 
         path = plot_dir, 
         width = 10, height = 8, 
         device = "pdf", dpi = 600)
  
  p4
}
```

## 8. Excitatory Module Violin Plot

```{r}
# Check if one_over_f_status exists
if (!"one_over_f_status" %in% colnames(neuron_obj_cleaned@meta.data)) {
  warning("one_over_f_status not found in metadata. Skipping violin plot.")
} else {
  # Violin plot of Excitatory module score
  p5 <- VlnPlot(neuron_obj_cleaned, 
                features = c("Excitatory_Profile1"), 
                group.by = "one_over_f_status") + 
    ggtitle("Excitatory Profile: Steep vs Flat 1/f") + 
    ylab("Excitatory Module Score") +
    theme(plot.title = element_text(hjust = 0.5))
  
  ggsave(filename = "epilepsy_excitatory_module_violin.pdf", 
         plot = p5, 
         path = plot_dir, 
         width = 8, height = 6, 
         device = "pdf", dpi = 600)
  
  p5
  
  # Also create violin plot by subclass
  p6 <- VlnPlot(neuron_obj_cleaned, 
                features = c("Excitatory_Profile1"), 
                group.by = "subclass_name") + 
    ggtitle("Excitatory Profile by Neuron Subtype") + 
    ylab("Excitatory Module Score") +
    theme(plot.title = element_text(hjust = 0.5),
          axis.text.x = element_text(angle = 45, hjust = 1))
  
  ggsave(filename = "epilepsy_excitatory_module_by_subtype.pdf", 
         plot = p6, 
         path = plot_dir, 
         width = 12, height = 6, 
         device = "pdf", dpi = 600)
  
  p6
}
```

## 9. Export Data for Prism

```{r}
# Create CSV output directory
csv_dir <- "D:/AbrahamDada/oneoverf/oneoverf_R1_10326/csv"
if (!dir.exists(csv_dir)) {
  dir.create(csv_dir, recursive = TRUE)
}

# Extract metadata with module scores
prism_data <- neuron_obj_cleaned@meta.data

# Save as CSV
csv_path <- file.path(csv_dir, "epilepsy_module_scores.csv")
write.csv(prism_data, csv_path, row.names = TRUE)

cat("Data exported to:", csv_path, "\n")
cat("Columns exported:", paste(colnames(prism_data), collapse = ", "), "\n")
```

## 10. Save Annotated Objects

```{r}
# Save the full object with MapMyCells annotations
output_path <- "D:/Youssef Sibih/Seurat_Objects/R1_123125/epilepsy_4samples_combined_sct_MapCells.rds"
saveRDS(obj, file = output_path)
cat("Full annotated object saved to:", output_path, "\n")

# Save the neuron-only object with MapMyCells annotations and module scores
neuron_output_path <- "D:/Youssef Sibih/Seurat_Objects/R1_123125/epilepsy_neurons_sct_MapCells.rds"
saveRDS(neuron_obj_cleaned, file = neuron_output_path)
cat("Neuron-only object saved to:", neuron_output_path, "\n")
```

## 11. Summary

```{r}
cat("=== Analysis Summary ===\n")
cat("Total cells in object:", ncol(obj), "\n")
cat("Cells with MapMyCells annotations:", sum(!is.na(obj@meta.data$class_name)), "\n")
cat("Total neurons:", ncol(neuron_obj_cleaned), "\n")
cat("Neuron subtypes identified:", length(unique(neuron_obj_cleaned@meta.data$subclass_name)), "\n\n")

cat("Outputs created:\n")
cat("1. Full annotated object:", output_path, "\n")
cat("2. Neuron-only object:", neuron_output_path, "\n")
cat("3. Plots saved to:", plot_dir, "\n")
cat("   - epilepsy_allcells_labeled.pdf\n")
cat("   - epilepsy_allcells_unlabeled.pdf\n")
cat("   - epilepsy_neurononly_labeled.pdf\n")
cat("   - epilepsy_neurononly_nolabel.pdf\n")
if ("one_over_f_status" %in% colnames(neuron_obj_cleaned@meta.data)) {
  cat("   - epilepsy_neurononly_1overf_labeled.pdf\n")
  cat("   - epilepsy_neurononly_1overf_nolabel.pdf\n")
  cat("   - epilepsy_proportions_1overf.pdf\n")
  cat("   - epilepsy_excitatory_module_violin.pdf\n")
  cat("   - epilepsy_excitatory_module_by_subtype.pdf\n")
}
cat("4. CSV export:", csv_path, "\n")
```

