---
title: "Neftel Cell States Analysis"
output: html_notebook
---

This script assigns Neftel glioma cell states (AC, MES, NPC, OPC) to tumor neurons and compares their proportions in steep vs flat 1/f across different glioma types.

## 1. Load Libraries

```{r}
library(Seurat)
library(dplyr)
library(ggplot2)
library(readxl)
library(tidyr)
```

## 2. Load and Process Neftel Gene Lists

```{r}
# Read the Neftel gene lists from Excel file (Table S2)
# Skip first 3 rows (description), row 4 has headers, data starts row 5
excel_path <- "C:/Users/Administrator/Downloads/1-s2.0-S0092867419306877-mmc2.xlsx"

cat("Reading Excel file...\n")
cat("File path:", excel_path, "\n\n")

# Read starting from row 4 (skip first 3 description rows)
gene_lists_raw <- read_excel(excel_path, sheet = 1, skip = 3)

cat("Gene lists loaded from Excel\n")
cat("Dimensions:", nrow(gene_lists_raw), "rows x", ncol(gene_lists_raw), "columns\n")
cat("Column names:", paste(colnames(gene_lists_raw), collapse = ", "), "\n\n")

# Display first few rows
cat("First 10 genes from each list:\n")
print(head(gene_lists_raw, 10))

# Extract gene lists for each state
# Column names should now be: MES2, MES1, AC, OPC, NPC1, NPC2, G1/S, G2/M
MES1_genes <- na.omit(as.character(gene_lists_raw$MES1))
MES2_genes <- na.omit(as.character(gene_lists_raw$MES2))
AC_genes <- na.omit(as.character(gene_lists_raw$AC))
OPC_genes <- na.omit(as.character(gene_lists_raw$OPC))
NPC1_genes <- na.omit(as.character(gene_lists_raw$NPC1))
NPC2_genes <- na.omit(as.character(gene_lists_raw$NPC2))

cat("Gene list sizes:\n")
cat("  MES1:", length(MES1_genes), "genes\n")
cat("  MES2:", length(MES2_genes), "genes\n")
cat("  AC:", length(AC_genes), "genes\n")
cat("  OPC:", length(OPC_genes), "genes\n")
cat("  NPC1:", length(NPC1_genes), "genes\n")
cat("  NPC2:", length(NPC2_genes), "genes\n")

# Save gene lists as CSV for reference
csv_dir <- "D:/AbrahamDada/oneoverf/oneoverf_R1_10326/csv"
if (!dir.exists(csv_dir)) {
  dir.create(csv_dir, recursive = TRUE)
}

write.csv(gene_lists_raw, file.path(csv_dir, "neftel_gene_lists.csv"), row.names = FALSE)
cat("\nGene lists saved to:", file.path(csv_dir, "neftel_gene_lists.csv"), "\n")
```

## 3. Load Neurons + Tumor Object and Subset to Tumor Cells

```{r}
# Load the neurons + tumor object
obj_full <- readRDS("D:/Youssef Sibih/Seurat_Objects/neurons_tumor_Glioma.rds")

cat("Object loaded successfully!\n")
cat("Total cells:", ncol(obj_full), "\n")

# Check what column identifies tumor cells
cat("\nChecking for tumor cell identifier...\n")
cat("Metadata columns:", paste(colnames(obj_full@meta.data), collapse = ", "), "\n\n")

# Check Idents
cat("Current Idents:\n")
print(table(Idents(obj_full)))

# Check manual_annot if it exists
if ("manual_annot" %in% colnames(obj_full@meta.data)) {
  cat("\nmanual_annot values:\n")
  print(table(obj_full$manual_annot, useNA = "ifany"))
}

# Subset to Tumor cells only
# Assuming Idents or manual_annot contains "Tumor" for tumor cells
if ("manual_annot" %in% colnames(obj_full@meta.data)) {
  obj <- subset(obj_full, subset = manual_annot == "Tumor")
  cat("\n✓ Subset to Tumor cells using manual_annot\n")
} else if ("manual_anot" %in% colnames(obj_full@meta.data)) {
  obj <- subset(obj_full, subset = manual_anot == "Tumor")
  cat("\n✓ Subset to Tumor cells using manual_anot\n")
} else {
  # Try using Idents if manual_annot doesn't exist
  Idents(obj_full) <- obj_full@active.ident
  obj <- subset(obj_full, idents = "Tumor")
  cat("\n✓ Subset to Tumor cells using Idents\n")
}

cat("Tumor cells only:", ncol(obj), "\n")
cat("Unique cortex types in tumor cells:", paste(unique(obj$cortex_type), collapse = ", "), "\n\n")

# Display distribution by cortex type
cat("Tumor cell distribution by cortex type:\n")
print(table(obj$cortex_type, useNA = "ifany"))

# Display 1/f status distribution in tumor cells
if ("one_over_f_status" %in% colnames(obj@meta.data)) {
  cat("\nTumor cell distribution by one_over_f_status:\n")
  print(table(obj$one_over_f_status, useNA = "ifany"))
  
  cat("\nCross-tabulation (cortex_type x one_over_f_status):\n")
  print(table(obj$cortex_type, obj$one_over_f_status, useNA = "ifany"))
}

# Basic UMAP
DimPlot(obj, group.by = "cortex_type", label = TRUE) +
  ggtitle("Tumor Cells Only")
```

## 4. Calculate Neftel Module Scores

```{r}
cat("Calculating module scores for Neftel cell states...\n")

# Add module scores for each state
# Following Neftel et al. methods: assay="SCT", nbin=30, search=TRUE
obj <- AddModuleScore(obj, features = list(MES1_genes), 
                      name = "MES1_score", assay = "SCT", nbin = 30, search = TRUE)

obj <- AddModuleScore(obj, features = list(MES2_genes), 
                      name = "MES2_score", assay = "SCT", nbin = 30, search = TRUE)

obj <- AddModuleScore(obj, features = list(AC_genes), 
                      name = "AC_score", assay = "SCT", nbin = 30, search = TRUE)

obj <- AddModuleScore(obj, features = list(OPC_genes), 
                      name = "OPC_score", assay = "SCT", nbin = 30, search = TRUE)

obj <- AddModuleScore(obj, features = list(NPC1_genes), 
                      name = "NPC1_score", assay = "SCT", nbin = 30, search = TRUE)

obj <- AddModuleScore(obj, features = list(NPC2_genes), 
                      name = "NPC2_score", assay = "SCT", nbin = 30, search = TRUE)

cat("Module scores calculated successfully!\n")
cat("New columns added:", "MES1_score1, MES2_score1, AC_score1, OPC_score1, NPC1_score1, NPC2_score1\n")
```

## 5. Assign Cell States

```{r}
cat("Assigning cell states based on highest module score...\n")

# Create a matrix of scores for comparison
score_matrix <- data.frame(
  MES1 = obj$MES1_score1,
  MES2 = obj$MES2_score1,
  AC = obj$AC_score1,
  OPC = obj$OPC_score1,
  NPC1 = obj$NPC1_score1,
  NPC2 = obj$NPC2_score1
)

# Find which state has the highest score for each cell
highest_state <- apply(score_matrix, 1, function(x) colnames(score_matrix)[which.max(x)])

# Collapse MES1/MES2 to MES, and NPC1/NPC2 to NPC
neftel_state <- case_when(
  highest_state == "MES1" ~ "MES",
  highest_state == "MES2" ~ "MES",
  highest_state == "NPC1" ~ "NPC",
  highest_state == "NPC2" ~ "NPC",
  highest_state == "AC" ~ "AC",
  highest_state == "OPC" ~ "OPC",
  TRUE ~ NA_character_
)

# Add to metadata
obj$neftel_state <- neftel_state

cat("Cell state assignment complete!\n")
cat("State distribution:\n")
print(table(obj$neftel_state, useNA = "ifany"))

# Visualize
DimPlot(obj, group.by = "neftel_state", label = TRUE) +
  ggtitle("Neftel Cell States")
```

## 6. Generate Stacked Bar Plot (All in One Figure)

```{r}
library(patchwork)

# Define color scheme for cell states
neftel_colors <- c(
  "AC" = "#FF7F00",    # Orange
  "MES" = "#E377C2",   # Pink
  "NPC" = "#4DAF4A",   # Green
  "OPC" = "#8C564B"    # Brown
)

# Create output directory for plots
plot_dir <- "D:/AbrahamDada/oneoverf/oneoverf_R1_10326/plots"
if (!dir.exists(plot_dir)) {
  dir.create(plot_dir, recursive = TRUE)
}

# Create combined proportion table for all conditions
prop_data <- obj@meta.data %>%
  group_by(cortex_type, one_over_f_status, neftel_state) %>%
  summarise(Freq = n(), .groups = "drop") %>%
  group_by(cortex_type, one_over_f_status) %>%
  mutate(prop = Freq / sum(Freq)) %>%
  ungroup()

# Create factors with proper ordering
prop_data <- prop_data %>%
  mutate(
    cortex_type = factor(cortex_type, levels = c("oligodendroglioma", "astrocytoma", "glioblastoma")),
    one_over_f_status = factor(one_over_f_status, levels = c("high1overf", "low1overf")),
    neftel_state = factor(neftel_state, levels = c("AC", "MES", "NPC", "OPC")),
    # Create short labels for x-axis
    status_label = case_when(
      one_over_f_status == "high1overf" ~ "High",
      one_over_f_status == "low1overf" ~ "Low"
    )
  )

# Create stacked bar plot with faceting for visual grouping
p_bars <- ggplot(prop_data, aes(x = status_label, y = prop, fill = neftel_state)) +
  geom_bar(stat = "identity", position = "stack", width = 0.6) +
  facet_wrap(~ cortex_type, nrow = 1, strip.position = "bottom",
             labeller = labeller(cortex_type = c(
               "oligodendroglioma" = "Oligodendroglioma",
               "astrocytoma" = "Astrocytoma",
               "glioblastoma" = "Glioblastoma"
             ))) +
  scale_fill_manual(values = neftel_colors, name = "Cell State") +
  labs(x = NULL, y = "Proportion") +
  theme_classic(base_size = 11) +
  theme(axis.text.x = element_text(size = 9),
        strip.background = element_rect(fill = "white", color = "black"),
        strip.text = element_text(face = "bold", size = 10),
        strip.placement = "outside",
        legend.position = "right",
        panel.spacing = unit(1.5, "lines"))  # Add spacing between facets

# Save the bar plot
ggsave(filename = "neftel_states_barplot.pdf", 
       plot = p_bars, 
       path = plot_dir, 
       width = 10, height = 6, 
       device = "pdf", dpi = 600)

p_bars
```

## 7. Generate Quadrant Plots (3x2 Grid)

```{r}
# Calculate coordinates for quadrant plots following Neftel et al. methodology

# For MES scores, use the maximum of MES1 and MES2
obj$MES_score <- pmax(obj$MES1_score1, obj$MES2_score1)

# For NPC scores, use the maximum of NPC1 and NPC2
obj$NPC_score <- pmax(obj$NPC1_score1, obj$NPC2_score1)

# Calculate x and y coordinates
obj$neftel_x <- case_when(
  obj$neftel_state %in% c("NPC", "OPC") ~ log2(abs(obj$NPC_score - obj$OPC_score1) + 1),
  obj$neftel_state %in% c("AC", "MES") ~ log2(abs(obj$MES_score - obj$AC_score1) + 1),
  TRUE ~ 0
)

# Multiply x by -1 for OPC and AC (to place them on the left)
obj$neftel_x <- ifelse(obj$neftel_state %in% c("OPC", "AC"), 
                       -obj$neftel_x, 
                       obj$neftel_x)

# Calculate y coordinate (same for all cells)
obj$neftel_y <- pmax(obj$OPC_score1, obj$NPC_score) - pmax(obj$AC_score1, obj$MES_score)

# Function to create quadrant plot for one condition
plot_quadrant <- function(obj, cortex_type_name, oneoverf_status, colors) {
  # Subset
  obj_subset <- subset(obj, subset = cortex_type == cortex_type_name & one_over_f_status == oneoverf_status)
  
  # Extract data
  plot_df <- data.frame(
    x = obj_subset$neftel_x,
    y = obj_subset$neftel_y,
    state = obj_subset$neftel_state
  )
  
  # Create plot
  p <- ggplot(plot_df, aes(x = x, y = y, color = state)) +
    geom_point(size = 0.5, alpha = 0.6) +
    scale_color_manual(values = colors, name = "Cell State") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
    labs(title = paste0(cortex_type_name, " - ", oneoverf_status),
         x = "AC/OPC ← → MES/NPC",
         y = "OPC/NPC ← → AC/MES") +
    theme_classic(base_size = 9) +
    theme(plot.title = element_text(hjust = 0.5, size = 8),
          legend.position = "none")
  
  return(p)
}

# Generate all 6 quadrant plots
# Row 1: Oligodendroglioma (high, low)
q1 <- plot_quadrant(obj, "oligodendroglioma", "high1overf", neftel_colors)
q2 <- plot_quadrant(obj, "oligodendroglioma", "low1overf", neftel_colors)

# Row 2: Astrocytoma (high, low)
q3 <- plot_quadrant(obj, "astrocytoma", "high1overf", neftel_colors)
q4 <- plot_quadrant(obj, "astrocytoma", "low1overf", neftel_colors)

# Row 3: Glioblastoma (high, low)
q5 <- plot_quadrant(obj, "glioblastoma", "high1overf", neftel_colors)
q6 <- plot_quadrant(obj, "glioblastoma", "low1overf", neftel_colors)

# Combine into 3x2 grid
combined_quadrants <- (q1 | q2) / (q3 | q4) / (q5 | q6) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Neftel Cell States - Quadrant Plots",
    theme = theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"))
  )

# Save the quadrant plots
ggsave(filename = "neftel_states_quadrants.pdf", 
       plot = combined_quadrants, 
       path = plot_dir, 
       width = 10, height = 12, 
       device = "pdf", dpi = 600)

combined_quadrants
```

## 8. Chi-Square Analysis for Each Pathology

```{r}
library(gridExtra)
library(grid)

cat("=== Chi-Square Analysis ===\n\n")

# Function to perform chi-square test and create result table
chisq_analysis <- function(obj, cortex_type_name) {
  # Subset to specific cortex type
  obj_subset <- subset(obj, subset = cortex_type == cortex_type_name)
  
  # Create contingency table
  cont_table <- table(obj_subset$one_over_f_status, obj_subset$neftel_state)
  
  cat(cortex_type_name, ":\n")
  print(cont_table)
  cat("\n")
  
  # Perform chi-square test
  chisq_result <- chisq.test(cont_table)
  
  cat("Chi-square statistic:", chisq_result$statistic, "\n")
  cat("P-value:", chisq_result$p.value, "\n")
  cat("Degrees of freedom:", chisq_result$parameter, "\n\n")
  
  # Create result data frame
  result_df <- data.frame(
    Statistic = c("Chi-square", "P-value", "Degrees of freedom", "Significant (p<0.05)"),
    Value = c(
      round(chisq_result$statistic, 4),
      ifelse(chisq_result$p.value < 2.2e-16, 
             "< 2.2e-16", 
             formatC(chisq_result$p.value, format = "e", digits = 10)),
      chisq_result$parameter,
      ifelse(chisq_result$p.value < 0.05, "Yes", "No")
    )
  )
  
  # Add contingency table to result with n (%) format
  cont_df <- as.data.frame.matrix(cont_table)
  
  # Calculate percentages for each row
  cont_df_pct <- cont_df / rowSums(cont_df) * 100
  
  # Combine counts and percentages as "n (%)"
  for (col in colnames(cont_df)) {
    cont_df[[col]] <- paste0(cont_df[[col]], " (", round(cont_df_pct[[col]], 1), "%)")
  }
  
  cont_df$one_over_f_status <- rownames(cont_df)
  cont_df <- cont_df[, c("one_over_f_status", colnames(cont_df)[1:(ncol(cont_df)-1)])]
  
  # Create a table plot
  title_grob <- textGrob(paste("Chi-Square Test:", cortex_type_name), 
                         gp = gpar(fontsize = 14, fontface = "bold"))
  
  subtitle_grob <- textGrob("Contingency Table (one_over_f_status x neftel_state)", 
                            gp = gpar(fontsize = 10))
  
  cont_table_grob <- tableGrob(cont_df, rows = NULL, 
                               theme = ttheme_default(base_size = 10))
  
  spacer <- rectGrob(gp = gpar(col = NA, fill = NA), height = unit(0.3, "cm"))
  
  stats_grob <- tableGrob(result_df, rows = NULL, 
                          theme = ttheme_default(base_size = 10))
  
  # Combine into one plot
  combined_grob <- arrangeGrob(title_grob, subtitle_grob, spacer, 
                               cont_table_grob, spacer, stats_grob,
                               ncol = 1, heights = c(0.5, 0.3, 0.2, 1.5, 0.2, 1))
  
  return(list(result = chisq_result, grob = combined_grob))
}

# Perform chi-square analysis for each cortex type
chisq_oligo <- chisq_analysis(obj, "oligodendroglioma")
chisq_astro <- chisq_analysis(obj, "astrocytoma")
chisq_gbm <- chisq_analysis(obj, "glioblastoma")

# Save each as PDF
ggsave(filename = "neftel_chisquare_oligodendroglioma.pdf", 
       plot = chisq_oligo$grob, 
       path = plot_dir, 
       width = 8, height = 6, 
       device = "pdf", dpi = 600)

ggsave(filename = "neftel_chisquare_astrocytoma.pdf", 
       plot = chisq_astro$grob, 
       path = plot_dir, 
       width = 8, height = 6, 
       device = "pdf", dpi = 600)

ggsave(filename = "neftel_chisquare_glioblastoma.pdf", 
       plot = chisq_gbm$grob, 
       path = plot_dir, 
       width = 8, height = 6, 
       device = "pdf", dpi = 600)

cat("Chi-square test PDFs saved for each pathology\n")
```

## 9. Summary Statistics

```{r}
cat("=== Neftel Cell States Summary ===\n\n")

# Overall distribution
cat("Overall state distribution:\n")
print(table(obj$neftel_state, useNA = "ifany"))

# By cortex type
cat("\nBy cortex type:\n")
print(table(obj$cortex_type, obj$neftel_state, useNA = "ifany"))

# By one_over_f_status
if ("one_over_f_status" %in% colnames(obj@meta.data)) {
  cat("\nBy one_over_f_status:\n")
  print(table(obj$one_over_f_status, obj$neftel_state, useNA = "ifany"))
  
  # Three-way table
  cat("\nDetailed breakdown (cortex_type x one_over_f_status x neftel_state):\n")
  for (ct in unique(obj$cortex_type)) {
    cat("\n", ct, ":\n")
    obj_ct <- subset(obj, subset = cortex_type == ct)
    print(table(obj_ct$one_over_f_status, obj_ct$neftel_state, useNA = "ifany"))
  }
}

cat("\n=== Outputs Created ===\n")
cat("1. Gene lists CSV:", file.path(csv_dir, "neftel_gene_lists.csv"), "\n")
cat("2. Plots saved to:", plot_dir, "\n")
cat("   - neftel_states_barplot.pdf (6 bars: high/low for oligo, astro, gbm)\n")
cat("   - neftel_states_quadrants.pdf (3x2 grid: quadrant plots for each condition)\n")
cat("   - neftel_chisquare_oligodendroglioma.pdf (chi-square test table)\n")
cat("   - neftel_chisquare_astrocytoma.pdf (chi-square test table)\n")
cat("   - neftel_chisquare_glioblastoma.pdf (chi-square test table)\n")
```

