---
title: "Comparative DGE Analysis: Epilepsy vs Normal vs Tumor"
output: html_notebook
---

This script performs differential gene expression analysis on normal and tumor neurons, then compares results with epilepsy to identify shared and unique DEGs.

## 1. Load Libraries

```{r}
library("clusterProfiler")
library("AnnotationDbi")
library("org.Hs.eg.db")
library(EnhancedVolcano)
library(writexl)
library(Seurat)
library(dplyr)
library(ggplot2)
library(VennDiagram)
library(gridExtra)
```

## 2. Load Epilepsy DGE Results

```{r}
# Load the epilepsy DGE results from previous analysis
epilepsy_dge <- read.csv("D:/AbrahamDada/oneoverf/oneoverf_R1_10326/csv/epilepsy_DGE_oneoverf.csv")

cat("Epilepsy DGE results loaded\n")
cat("Total genes:", nrow(epilepsy_dge), "\n")

# Define significant DEGs (threshold: |log2FC| > 0.25 & p_adj < 0.05)
epilepsy_degs_high <- epilepsy_dge$Gene[epilepsy_dge$avg_log2FC < -0.25 & epilepsy_dge$p_val_adj < 0.05]
epilepsy_degs_low <- epilepsy_dge$Gene[epilepsy_dge$avg_log2FC > 0.25 & epilepsy_dge$p_val_adj < 0.05]

cat("Epilepsy DEGs (steep 1/f):", length(epilepsy_degs_high), "\n")
cat("Epilepsy DEGs (flat 1/f):", length(epilepsy_degs_low), "\n")
```

## 3. Load Main Neuron Object and Prepare for DGE

```{r}
# Load the prepared main neuron object
main_neuron_obj <- readRDS("D:/Youssef Sibih/Seurat_Objects/mainobj_metadata_corrected_neurons_only.rds")

cat("Main neuron object loaded\n")
cat("Total cells before filtering:", ncol(main_neuron_obj), "\n")

# Check class_name distribution
cat("\nclass_name distribution:\n")
print(table(main_neuron_obj$class_name, useNA = "ifany"))

# Filter to only neuronal populations (exclude "Non-neuronal and Non-neural")
cat("\nFiltering to neuronal populations only...\n")
main_neuron_obj <- subset(main_neuron_obj, 
                          subset = class_name %in% c("Neuronal: Glutamatergic", "Neuronal: GABAergic"))

cat("Filtered to neuronal populations\n")
cat("Total neurons after filtering:", ncol(main_neuron_obj), "\n")
cat("Cortex types:", paste(unique(main_neuron_obj$cortex_type), collapse = ", "), "\n\n")

# Display distribution
cat("Distribution by cortex_type:\n")
print(table(main_neuron_obj$cortex_type, useNA = "ifany"))

cat("\nDistribution by one_over_f_status:\n")
print(table(main_neuron_obj$one_over_f_status, useNA = "ifany"))

cat("\nCross-tabulation:\n")
print(table(main_neuron_obj$cortex_type, main_neuron_obj$one_over_f_status, useNA = "ifany"))

# IMPORTANT: Prepare the FULL filtered object for SCT-based FindMarkers BEFORE further subsetting
cat("\nPreparing full filtered object for SCT-based DGE...\n")
main_neuron_obj <- PrepSCTFindMarkers(main_neuron_obj)
cat("PrepSCTFindMarkers complete on full filtered object\n")
```

## 4. DGE Analysis - Normal Neurons

```{r}
# Subset to Normal Cortex neurons (AFTER PrepSCTFindMarkers)
normal_obj <- subset(main_neuron_obj, subset = cortex_type == "Normal Cortex")

cat("Normal neurons subset created\n")
cat("Total cells:", ncol(normal_obj), "\n")
cat("one_over_f_status distribution:\n")
print(table(normal_obj$one_over_f_status))

# Run DGE (object already prepared, use recorrect_umi=FALSE)
Idents(normal_obj) <- "one_over_f_status"

normal_dge <- FindMarkers(normal_obj, 
                          ident.1 = "low1overf", 
                          ident.2 = "high1overf",
                          verbose = FALSE,
                          recorrect_umi = FALSE)

# Add gene names as column
normal_result <- cbind(Gene = rownames(normal_dge), normal_dge)

# Save CSV
csv_dir <- "D:/AbrahamDada/oneoverf/oneoverf_R1_10326/csv"
write.csv(normal_result, file.path(csv_dir, "normal_DGE_oneoverf.csv"), row.names = FALSE)

cat("Normal DGE complete. Saved to:", file.path(csv_dir, "normal_DGE_oneoverf.csv"), "\n")

# Define significant DEGs
normal_degs_high <- normal_result$Gene[normal_result$avg_log2FC < -0.25 & normal_result$p_val_adj < 0.05]
normal_degs_low <- normal_result$Gene[normal_result$avg_log2FC > 0.25 & normal_result$p_val_adj < 0.05]

cat("Normal DEGs (steep 1/f):", length(normal_degs_high), "\n")
cat("Normal DEGs (flat 1/f):", length(normal_degs_low), "\n")
```

## 5. DGE Analysis - All Tumor Neurons

```{r}
# Subset to all tumor neurons (AFTER PrepSCTFindMarkers)
tumor_obj <- subset(main_neuron_obj, subset = cortex_type %in% c("oligodendroglioma", "astrocytoma", "glioblastoma"))

cat("Tumor neurons subset created\n")
cat("Total cells:", ncol(tumor_obj), "\n")
cat("Cortex types:", paste(unique(tumor_obj$cortex_type), collapse = ", "), "\n")

# Run DGE on all tumor cells (object already prepared, use recorrect_umi=FALSE)
Idents(tumor_obj) <- "one_over_f_status"

tumor_all_dge <- FindMarkers(tumor_obj, 
                             ident.1 = "low1overf", 
                             ident.2 = "high1overf",
                             verbose = FALSE,
                             recorrect_umi = FALSE)

# Add gene names as column
tumor_all_result <- cbind(Gene = rownames(tumor_all_dge), tumor_all_dge)

# Save CSV
write.csv(tumor_all_result, file.path(csv_dir, "tumor_all_DGE_oneoverf.csv"), row.names = FALSE)

cat("Tumor (all) DGE complete. Saved to:", file.path(csv_dir, "tumor_all_DGE_oneoverf.csv"), "\n")

# Define significant DEGs
tumor_all_degs_high <- tumor_all_result$Gene[tumor_all_result$avg_log2FC < -0.25 & tumor_all_result$p_val_adj < 0.05]
tumor_all_degs_low <- tumor_all_result$Gene[tumor_all_result$avg_log2FC > 0.25 & tumor_all_result$p_val_adj < 0.05]

cat("Tumor (all) DEGs (steep 1/f):", length(tumor_all_degs_high), "\n")
cat("Tumor (all) DEGs (flat 1/f):", length(tumor_all_degs_low), "\n")
```

## 6. DGE Analysis - Tumor Subtypes

```{r}
# Get unique cortex types
types <- c("oligodendroglioma", "astrocytoma", "glioblastoma")

# Initialize list to store results
subtype_results <- list()
subtype_degs_high <- list()
subtype_degs_low <- list()

for(ct in types) {
  cat("\n=== Processing:", ct, "===\n")
  
  # Subset to current cortex_type
  sub_obj <- subset(tumor_obj, subset = cortex_type == ct)
  cat("Cells:", ncol(sub_obj), "\n")
  
  # Perform differential expression analysis
  result <- FindMarkers(
    sub_obj, 
    ident.1 = "low1overf",
    ident.2 = "high1overf",
    group.by = "one_over_f_status",
    logfc.threshold = 0,
    min.pct = 0,
    verbose = FALSE,
    recorrect_umi = FALSE
  )
  
  # Genes are stored as rownames, moving them to their own column
  result <- cbind(Gene = rownames(result), result)
  
  # Save as CSV
  csv_filename <- paste0("tumor_", ct, "_DGE_oneoverf.csv")
  write.csv(result, file.path(csv_dir, csv_filename), row.names = FALSE)
  cat("Saved:", csv_filename, "\n")
  
  # Store results
  subtype_results[[ct]] <- result
  
  # Define significant DEGs
  subtype_degs_high[[ct]] <- result$Gene[result$avg_log2FC < -0.25 & result$p_val_adj < 0.05]
  subtype_degs_low[[ct]] <- result$Gene[result$avg_log2FC > 0.25 & result$p_val_adj < 0.05]
  
  cat("DEGs (steep 1/f):", length(subtype_degs_high[[ct]]), "\n")
  cat("DEGs (flat 1/f):", length(subtype_degs_low[[ct]]), "\n")
}
```

## 6. Overlap Analysis - Bar Plots

```{r}
# Calculate overlaps
calc_overlap_pct <- function(set1, set2) {
  overlap <- length(intersect(set1, set2))
  pct <- (overlap / length(set1)) * 100
  return(list(overlap = overlap, pct = pct, total = length(set1)))
}

# Calculate overlaps for steep 1/f
overlap_tumor_high <- calc_overlap_pct(epilepsy_degs_high, tumor_all_degs_high)
overlap_normal_high <- calc_overlap_pct(epilepsy_degs_high, normal_degs_high)

# Calculate overlaps for flat 1/f
overlap_tumor_low <- calc_overlap_pct(epilepsy_degs_low, tumor_all_degs_low)
overlap_normal_low <- calc_overlap_pct(epilepsy_degs_low, normal_degs_low)

# Create data frame for plotting
overlap_df <- data.frame(
  Comparison = factor(c("Epilepsy vs Tumor\n(Steep 1/f)", 
                       "Epilepsy vs Tumor\n(Flat 1/f)",
                       "Epilepsy vs Normal\n(Steep 1/f)", 
                       "Epilepsy vs Normal\n(Flat 1/f)"),
                     levels = c("Epilepsy vs Tumor\n(Steep 1/f)", 
                               "Epilepsy vs Tumor\n(Flat 1/f)",
                               "Epilepsy vs Normal\n(Steep 1/f)", 
                               "Epilepsy vs Normal\n(Flat 1/f)")),
  Overlap_Pct = c(overlap_tumor_high$pct, overlap_tumor_low$pct,
                  overlap_normal_high$pct, overlap_normal_low$pct),
  Overlap_N = c(overlap_tumor_high$overlap, overlap_tumor_low$overlap,
                overlap_normal_high$overlap, overlap_normal_low$overlap),
  Total_N = c(overlap_tumor_high$total, overlap_tumor_low$total,
              overlap_normal_high$total, overlap_normal_low$total)
)

# Create bar plot with better aesthetics
plot_dir <- "D:/AbrahamDada/oneoverf/oneoverf_R1_10326/plots"

# Version 1: Colored bars
overlap_colors <- c("#E377C2", "#FF7F00", "#4DAF4A", "#8C564B")

p_overlap_colored <- ggplot(overlap_df, aes(x = Comparison, y = Overlap_Pct, fill = Comparison)) +
  geom_bar(stat = "identity", width = 0.75, colour = "black", size = 0.5) +
  geom_text(aes(label = paste0(round(Overlap_Pct, 1), "%\n(", Overlap_N, "/", Total_N, ")")),
            vjust = -0.3, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = overlap_colors, guide = "none") +
  labs(x = NULL, y = "% Overlap of DEGs") +
  theme_classic(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        axis.title.y = element_text(face = "bold", size = 12),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  ylim(0, max(overlap_df$Overlap_Pct) * 1.15)

ggsave(filename = "DGE_overlap_epilepsy_comparison_colored.pdf", 
       plot = p_overlap_colored, 
       path = plot_dir, 
       width = 5.2, height = 7.56, 
       device = "pdf", dpi = 600)

p_overlap_colored

# Version 2: All dark blue bars
p_overlap_blue <- ggplot(overlap_df, aes(x = Comparison, y = Overlap_Pct)) +
  geom_bar(stat = "identity", width = 0.75, fill = "#1f77b4", colour = "black", size = 0.5) +
  geom_text(aes(label = paste0(round(Overlap_Pct, 1), "%\n(", Overlap_N, "/", Total_N, ")")),
            vjust = -0.3, size = 3.5, fontface = "bold") +
  labs(x = NULL, y = "% Overlap of DEGs") +
  theme_classic(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        axis.title.y = element_text(face = "bold", size = 12),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  ylim(0, max(overlap_df$Overlap_Pct) * 1.15)

ggsave(filename = "DGE_overlap_epilepsy_comparison_blue.pdf", 
       plot = p_overlap_blue, 
       path = plot_dir, 
       width = 5.2, height = 7.56, 
       device = "pdf", dpi = 600)

p_overlap_blue
```

## 7. GO Enrichment - Tumor-Specific Genes (Not in Epilepsy)

```{r}
cat("\n=== GO Enrichment: Tumor-Specific Genes ===\n")

# Function to run GO enrichment for all ontologies
run_go_all_ontologies <- function(genes, title_prefix, file_prefix) {
  if (length(genes) < 10) {
    cat("  Not enough genes (", length(genes), ") for GO enrichment\n")
    return(NULL)
  }
  
  ontologies <- c("MF", "CC", "BP", "ALL")
  ont_names <- c("MF" = "Molecular Function", "CC" = "Cellular Component", 
                 "BP" = "Biological Process", "ALL" = "All Ontologies")
  
  for (ont in ontologies) {
    Go_result <- enrichGO(gene = genes, 
                         OrgDb = "org.Hs.eg.db", 
                         keyType = "SYMBOL", 
                         ont = ont, 
                         pvalueCutoff = 0.05, 
                         qvalueCutoff = 0.05)
    
    if (nrow(Go_result@result) > 0) {
      p <- dotplot(Go_result, showCategory = 10, orderBy = "GeneRatio") +
        ggtitle(paste0(title_prefix, " - ", ont_names[ont]))
      
      ggsave(filename = paste0(file_prefix, "_", ont, ".pdf"), 
             plot = p, path = plot_dir, 
             width = 10, height = 8, device = "pdf", dpi = 600)
      
      cat("  Saved:", paste0(file_prefix, "_", ont, ".pdf"), "\n")
    }
  }
}

# Genes unique to tumor (steep 1/f)
tumor_unique_high <- setdiff(tumor_all_degs_high, epilepsy_degs_high)
cat("Tumor-specific (steep 1/f):", length(tumor_unique_high), "genes\n")
run_go_all_ontologies(tumor_unique_high, 
                      "GO: Tumor-Specific (Steep 1/f)", 
                      "GO_tumor_unique_high1overf")

# Genes unique to tumor (flat 1/f)
tumor_unique_low <- setdiff(tumor_all_degs_low, epilepsy_degs_low)
cat("\nTumor-specific (flat 1/f):", length(tumor_unique_low), "genes\n")
run_go_all_ontologies(tumor_unique_low, 
                      "GO: Tumor-Specific (Flat 1/f)", 
                      "GO_tumor_unique_low1overf")
```

## 8. GO Enrichment - Epilepsy-Specific Genes (Not in Tumor)

```{r}
cat("\n=== GO Enrichment: Epilepsy-Specific Genes ===\n")

# Genes unique to epilepsy (steep 1/f)
epilepsy_unique_high <- setdiff(epilepsy_degs_high, tumor_all_degs_high)
cat("Epilepsy-specific (steep 1/f):", length(epilepsy_unique_high), "genes\n")
run_go_all_ontologies(epilepsy_unique_high, 
                      "GO: Epilepsy-Specific (Steep 1/f)", 
                      "GO_epilepsy_unique_high1overf")

# Genes unique to epilepsy (flat 1/f)
epilepsy_unique_low <- setdiff(epilepsy_degs_low, tumor_all_degs_low)
cat("\nEpilepsy-specific (flat 1/f):", length(epilepsy_unique_low), "genes\n")
run_go_all_ontologies(epilepsy_unique_low, 
                      "GO: Epilepsy-Specific (Flat 1/f)", 
                      "GO_epilepsy_unique_low1overf")
```

## 9. Subtype-Specific Comparisons

```{r}
cat("\n=== Subtype-Specific Comparisons ===\n")

for(ct in names(subtype_degs_high)) {
  cat("\n", ct, ":\n")
  
  # Calculate overlaps
  overlap_high <- calc_overlap_pct(epilepsy_degs_high, subtype_degs_high[[ct]])
  overlap_low <- calc_overlap_pct(epilepsy_degs_low, subtype_degs_low[[ct]])
  
  cat("  Overlap (steep 1/f):", overlap_high$pct, "% (", overlap_high$overlap, "/", overlap_high$total, ")\n")
  cat("  Overlap (flat 1/f):", overlap_low$pct, "% (", overlap_low$overlap, "/", overlap_low$total, ")\n")
  
  # GO enrichment for subtype-specific genes (not in epilepsy)
  # Steep 1/f
  subtype_unique_high <- setdiff(subtype_degs_high[[ct]], epilepsy_degs_high)
  cat("  ", ct, "-specific (steep 1/f):", length(subtype_unique_high), "genes\n")
  run_go_all_ontologies(subtype_unique_high, 
                        paste0("GO: ", ct, "-Specific (Steep 1/f)"), 
                        paste0("GO_", ct, "_unique_high1overf"))
  
  # Flat 1/f
  subtype_unique_low <- setdiff(subtype_degs_low[[ct]], epilepsy_degs_low)
  cat("  ", ct, "-specific (flat 1/f):", length(subtype_unique_low), "genes\n")
  run_go_all_ontologies(subtype_unique_low, 
                        paste0("GO: ", ct, "-Specific (Flat 1/f)"), 
                        paste0("GO_", ct, "_unique_low1overf"))
}
```

## 10. Export Gene Lists as CSVs

```{r}
cat("\n=== Exporting Gene Lists ===\n")

# HIGH 1/F GENE LISTS
# 1. Shared between epilepsy and all tumor
shared_epi_tumor_high <- intersect(epilepsy_degs_high, tumor_all_degs_high)
write.csv(data.frame(Gene = shared_epi_tumor_high), 
          file.path(csv_dir, "shared_epilepsy_tumor_high1overf.csv"), 
          row.names = FALSE)
cat("Saved: shared_epilepsy_tumor_high1overf.csv (", length(shared_epi_tumor_high), " genes)\n")

# 2. Unique to epilepsy
epilepsy_unique_high <- setdiff(epilepsy_degs_high, tumor_all_degs_high)
write.csv(data.frame(Gene = epilepsy_unique_high), 
          file.path(csv_dir, "unique_epilepsy_high1overf.csv"), 
          row.names = FALSE)
cat("Saved: unique_epilepsy_high1overf.csv (", length(epilepsy_unique_high), " genes)\n")

# 3. Unique to all tumor
tumor_unique_high <- setdiff(tumor_all_degs_high, epilepsy_degs_high)
write.csv(data.frame(Gene = tumor_unique_high), 
          file.path(csv_dir, "unique_tumor_high1overf.csv"), 
          row.names = FALSE)
cat("Saved: unique_tumor_high1overf.csv (", length(tumor_unique_high), " genes)\n")

# 4-6. Unique to each tumor subtype
for (ct in c("oligodendroglioma", "astrocytoma", "glioblastoma")) {
  subtype_unique <- setdiff(subtype_degs_high[[ct]], epilepsy_degs_high)
  write.csv(data.frame(Gene = subtype_unique), 
            file.path(csv_dir, paste0("unique_", ct, "_high1overf.csv")), 
            row.names = FALSE)
  cat("Saved: unique_", ct, "_high1overf.csv (", length(subtype_unique), " genes)\n")
}

cat("\n")

# LOW 1/F GENE LISTS
# 1. Shared between epilepsy and all tumor
shared_epi_tumor_low <- intersect(epilepsy_degs_low, tumor_all_degs_low)
write.csv(data.frame(Gene = shared_epi_tumor_low), 
          file.path(csv_dir, "shared_epilepsy_tumor_low1overf.csv"), 
          row.names = FALSE)
cat("Saved: shared_epilepsy_tumor_low1overf.csv (", length(shared_epi_tumor_low), " genes)\n")

# 2. Unique to epilepsy
epilepsy_unique_low <- setdiff(epilepsy_degs_low, tumor_all_degs_low)
write.csv(data.frame(Gene = epilepsy_unique_low), 
          file.path(csv_dir, "unique_epilepsy_low1overf.csv"), 
          row.names = FALSE)
cat("Saved: unique_epilepsy_low1overf.csv (", length(epilepsy_unique_low), " genes)\n")

# 3. Unique to all tumor
tumor_unique_low <- setdiff(tumor_all_degs_low, epilepsy_degs_low)
write.csv(data.frame(Gene = tumor_unique_low), 
          file.path(csv_dir, "unique_tumor_low1overf.csv"), 
          row.names = FALSE)
cat("Saved: unique_tumor_low1overf.csv (", length(tumor_unique_low), " genes)\n")

# 4-6. Unique to each tumor subtype
for (ct in c("oligodendroglioma", "astrocytoma", "glioblastoma")) {
  subtype_unique <- setdiff(subtype_degs_low[[ct]], epilepsy_degs_low)
  write.csv(data.frame(Gene = subtype_unique), 
            file.path(csv_dir, paste0("unique_", ct, "_low1overf.csv")), 
            row.names = FALSE)
  cat("Saved: unique_", ct, "_low1overf.csv (", length(subtype_unique), " genes)\n")
}

cat("\n=== Gene List Export Complete ===\n")
cat("Total CSV files created: 12 (6 for steep 1/f, 6 for flat 1/f)\n")
```

## 11. Summary

```{r}
cat("\n=== Comparative DGE Analysis Summary ===\n\n")

cat("Epilepsy DEGs:\n")
cat("  Steep 1/f:", length(epilepsy_degs_high), "\n")
  cat("  Flat 1/f:", length(epilepsy_degs_low), "\n\n")

if (exists("normal_degs_high")) {
  cat("Normal DEGs:\n")
  cat("  Steep 1/f:", length(normal_degs_high), "\n")
  cat("  Flat 1/f:", length(normal_degs_low), "\n\n")
}

cat("Tumor (all) DEGs:\n")
cat("  Steep 1/f:", length(tumor_all_degs_high), "\n")
cat("  Flat 1/f:", length(tumor_all_degs_low), "\n\n")

cat("Tumor Subtypes:\n")
for(ct in names(subtype_degs_high)) {
  cat("  ", ct, ":\n")
  cat("    Steep 1/f:", length(subtype_degs_high[[ct]]), "\n")
  cat("    Flat 1/f:", length(subtype_degs_low[[ct]]), "\n")
}

cat("\n=== Outputs Created ===\n")
cat("CSVs saved to:", csv_dir, "\n")
cat("  - DGE results for normal, tumor (all), and each subtype\n")
cat("  - Gene lists: shared and unique for steep/flat 1/f (12 CSV files)\n")
cat("\nPlots saved to:", plot_dir, "\n")
cat("  - DGE_overlap_epilepsy_comparison.pdf\n")
cat("  - GO enrichment plots (MF, CC, BP, ALL) for unique gene sets\n")
```

