---
title: "S0: Prepare Annotated Seurat Object"
output: html_notebook
---

This script prepares the cleaned, annotated Seurat object with:
- Corrected 1/f status
- Manual cell type annotations
- MapMyCells neuron annotations
- Patient metadata

## Setup

```{r setup}
library(dplyr)
library(Seurat)
library(ggplot2)
```

## 1. Load Raw Seurat Object

```{r load_object}
# Load the original Seurat object
obj <- readRDS("D:/Youssef Sibih/Seurat_Objects/oneoverf_combined_submap_cells_snc_sct_complete.rds")

# Quick view
print(paste("Total cells:", ncol(obj)))
DimPlot(obj, label = TRUE)
```

## 2. Fix 1/f Status Corrections

The original 1/f status assignments were incorrect for some samples. We need to fix:
- SF490_8 and SF495_6 should be high1overf (currently low1overf)
- SF490_3 and SF495_1 should be low1overf (currently high1overf)

```{r fix_1overf_status}
# Fix samples that should be high1overf
cells_to_update1 <- WhichCells(obj, 
                               expression = exp == "SF490_8" & one_over_f_status == "low1overf")
obj@meta.data[cells_to_update1, "one_over_f_status"] <- "high1overf"
print(paste("Fixed SF490_8:", length(cells_to_update1), "cells -> high1overf"))

cells_to_update2 <- WhichCells(obj, 
                               expression = exp == "SF495_6" & one_over_f_status == "low1overf")
obj@meta.data[cells_to_update2, "one_over_f_status"] <- "high1overf"
print(paste("Fixed SF495_6:", length(cells_to_update2), "cells -> high1overf"))

# Fix samples that should be low1overf
cells_to_update3 <- WhichCells(obj, 
                               expression = exp == "SF490_3" & one_over_f_status == "high1overf")
obj@meta.data[cells_to_update3, "one_over_f_status"] <- "low1overf"
print(paste("Fixed SF490_3:", length(cells_to_update3), "cells -> low1overf"))

cells_to_update4 <- WhichCells(obj, 
                               expression = exp == "SF495_1" & one_over_f_status == "high1overf")
obj@meta.data[cells_to_update4, "one_over_f_status"] <- "low1overf"
print(paste("Fixed SF495_1:", length(cells_to_update4), "cells -> low1overf"))

# Verify the fix
table(obj@meta.data$exp, obj@meta.data$one_over_f_status)
```

## 3. Add Manual Cell Type Annotations

Apply cluster annotations to assign cell types.

```{r manual_annotations}
# Store default cluster identities
obj@meta.data$default_ident <- Idents(obj)

# Define cluster to cell type mapping
cluster_ids <- c(
  "0" = "Tumor",
  "1" = "Oligodendrocyte",
  "2" = "Endothelial",
  "3" = "Oligodendrocyte",
  "4" = "Astrocyte",
  "5" = "Oligodendrocyte",
  "6" = "Neurons",
  "7" = "Myeloid/microglial",
  "8" = "Astrocyte",
  "9" = "Tumor",
  "10" = "Neurons",
  "11" = "Neurons",
  "12" = "Tumor",
  "13" = "Neurons",
  "14" = "Neurons",
  "15" = "Neurons",
  "16" = "Neurons",
  "17" = "Neurons",
  "18" = "Oligodendrocyte",
  "19" = "Tumor",
  "20" = "Oligodendrocyte",
  "21" = "Mesenchymal",
  "22" = "Erythroid",
  "23" = "Oligodendrocyte",
  "24" = "Neurons",
  "25" = "Tumor",
  "26" = "Myeloid",
  "27" = "Neurons",
  "28" = "Tumor",
  "29" = "Oligodendrocyte",
  "30" = "Tumor",
  "31" = "Neurons",
  "32" = "Neurons",
  "33" = "Astrocyte",
  "34" = "Tumor"
)

# Apply annotations
names(cluster_ids) <- levels(obj)
obj <- RenameIdents(obj, cluster_ids)

# Store in manual_anot column
obj$manual_anot <- Idents(obj)

# Verify annotations
print("Cell type counts:")
table(obj$manual_anot)

DimPlot(obj, label = TRUE) + ggtitle("Cell Type Annotations")
```

## 4. Add Cortex Type Metadata

```{r cortex_type}
# Initialize cortex_type with glioma_subtype values
obj@meta.data$cortex_type <- obj@meta.data$glioma_subtype

# Set Normal Cortex for control samples
obj@meta.data$cortex_type[obj@meta.data$exp %in% c("SF524_3", "SF524_1")] <- "Normal Cortex"

# Verify
print("Cortex type distribution:")
table(obj@meta.data$cortex_type)
```

## 5. Integrate MapMyCells Annotations

Add MapMyCells neuron subtype annotations (class_name and subclass_name) to the full object.
These will be NA for non-neurons.

```{r mapmycells_annotations}
# Define mapping files for Neurons groups
mapping_files <- list(
  neurons_1 = "C:/YS_Backup/Neurons_1_MTG/neurons_anndata_group1.csv",
  neurons_2 = "C:/YS_Backup/Neurons_2_MTG/neurons_anndata_group2.csv"
)

# Load and combine both annotations
all_mappings <- lapply(mapping_files, function(file) {
  mapping <- read.csv(file, comment.char = "#")
  rownames(mapping) <- mapping$cell_id
  return(mapping[, c("class_name", "subclass_name"), drop = FALSE])
})

# Merge both neuron group mappings
combined_mapping <- bind_rows(all_mappings)

# Find cells that exist in both the object and the mapping
common_cells <- intersect(rownames(obj@meta.data), rownames(combined_mapping))
print(paste("Neurons with MapMyCells annotations:", length(common_cells)))

# Initialize columns with NA for all cells
obj@meta.data$class_name <- NA
obj@meta.data$subclass_name <- NA

# Add annotations only for matching cells (neurons)
obj@meta.data[common_cells, "class_name"] <- combined_mapping[common_cells, "class_name"]
obj@meta.data[common_cells, "subclass_name"] <- combined_mapping[common_cells, "subclass_name"]

# Verify
print("Class name distribution (including NA for non-neurons):")
table(obj@meta.data$class_name, useNA = "ifany")

print("Subclass name distribution:")
table(obj@meta.data$subclass_name, useNA = "ifany")
```

## 6. Create Refined Annotations

Create `refined_annot` that combines manual annotations with MapMyCells neuron subtypes:
- Tumor cells keep "Tumor" designation
- Neurons with valid MapMyCells annotations (Glutamatergic or GABAergic) get their subclass_name
- All other cells keep their manual_anot

```{r refined_annotations}
# Start with manual_anot as base
obj@meta.data$refined_annot <- as.character(obj@meta.data$manual_anot)

# For neurons with valid neuronal MapMyCells annotations, use subclass_name
# Valid neuronal classes are "Neuronal: Glutamatergic" and "Neuronal: GABAergic"
valid_neuronal_classes <- c("Neuronal: Glutamatergic", "Neuronal: GABAergic")

# Find cells that are neurons AND have valid neuronal MapMyCells class
neurons_with_valid_mmc <- which(
  obj@meta.data$manual_anot == "Neurons" & 
  obj@meta.data$class_name %in% valid_neuronal_classes &
  !is.na(obj@meta.data$subclass_name)
)

# Update refined_annot for these cells with their subclass_name
obj@meta.data$refined_annot[neurons_with_valid_mmc] <- obj@meta.data$subclass_name[neurons_with_valid_mmc]

# Verify the results
print("Refined annotation distribution:")
print(table(obj@meta.data$refined_annot, useNA = "ifany"))

# Check that Tumor is preserved
print(paste("\nTumor cells in refined_annot:", sum(obj@meta.data$refined_annot == "Tumor")))
print(paste("Tumor cells in manual_anot:", sum(obj@meta.data$manual_anot == "Tumor")))

# Check neurons breakdown
print("\nNeurons breakdown:")
print(paste("  - Original 'Neurons' in manual_anot:", sum(obj@meta.data$manual_anot == "Neurons")))
print(paste("  - Neurons with valid MapMyCells subclass:", length(neurons_with_valid_mmc)))
print(paste("  - Neurons remaining as 'Neurons' in refined_annot:", sum(obj@meta.data$refined_annot == "Neurons")))
```

## 7. Add Patient Metadata

Extract patient ID from sample names (e.g., SF490_8 -> SF490).

```{r patient_metadata}
# Extract patient ID from exp field
obj@meta.data$patient_id <- gsub("_\\d+$", "", obj@meta.data$exp)

# Verify
print("Patient ID distribution:")
table(obj@meta.data$patient_id)

# Show sample to patient mapping
sample_patient_mapping <- unique(obj@meta.data[, c("exp", "patient_id", "one_over_f_status")])
sample_patient_mapping <- sample_patient_mapping[order(sample_patient_mapping$patient_id, 
                                                        sample_patient_mapping$one_over_f_status), ]
print("Sample to Patient mapping:")
print(sample_patient_mapping)
```

## 8. Final Verification

```{r final_verification}
# Summary of metadata columns
print("Metadata columns:")
print(colnames(obj@meta.data))

# Summary counts
print("Summary:")
print(paste("Total cells:", ncol(obj)))
print(paste("Samples:", length(unique(obj@meta.data$exp))))
print(paste("Patients:", length(unique(obj@meta.data$patient_id))))
print(paste("High 1/f cells:", sum(obj@meta.data$one_over_f_status == "high1overf")))
print(paste("Low 1/f cells:", sum(obj@meta.data$one_over_f_status == "low1overf")))
print(paste("Neurons with subclass annotations:", sum(!is.na(obj@meta.data$subclass_name))))

# View sample of metadata
head(obj@meta.data[, c("exp", "patient_id", "one_over_f_status", "manual_anot", "refined_annot", "class_name", "subclass_name", "cortex_type")])
```

## 9. Save Processed Object

Save the fully annotated object to Box.

```{r save_object}
# Define output path - update this to your Box sync folder location
output_path <- "D:/Youssef Sibih/Seurat_Objects/oneoverf_combined_corrected_annotations.rds"

# Save the object
saveRDS(obj, file = output_path)

print(paste("Object saved to:", output_path))
print(paste("File size:", round(file.info(output_path)$size / 1e9, 2), "GB"))
```

## Session Info

```{r session_info}
sessionInfo()
```
