---
title: "S1: Sample-Level UMAP Analysis"
output: html_notebook
---

This script generates sample-level UMAPs and cell count reports:
- Full UMAP colored by sample (all cells + neurons only)
- Grid of individual sample UMAPs organized by patient and 1/f status
- CSV reports of cell counts by sample and cell type

## Setup

```{r setup}
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
library(tidyr)
library(RColorBrewer)
```

## 1. Load Processed Object

```{r load_object}
# Load the annotated object created by S0
obj <- readRDS("D:/Youssef Sibih/Seurat_Objects/oneoverf_combined_corrected_annotations.rds")

# Create output directory if needed
output_dir <- "d:/AbrahamDada/oneoverf/oneoverf_R1_10326/plots"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# Quick verification
print(paste("Total cells:", ncol(obj)))
print("Samples:")
print(unique(obj@meta.data$exp))
```

## 2. Define Color Palettes

```{r color_palettes}
# Get unique samples and patients
samples <- sort(unique(obj@meta.data$exp))
patients <- sort(unique(obj@meta.data$patient_id))

# Create a highly distinguishable color palette for samples (14 samples)
# Combine colors from multiple palettes for maximum distinction
sample_colors <- c(
  "#E41A1C",
  "#377EB8",
  "#4DAF4A",
  "#984EA3",
  "#FF7F00",
  "#A65628",
  "#F781BF",
  "#1B9E77",
  "#D95F02",
  "#7570B3",
  "#E7298A",
  "#66A61E",
  "#E6AB02",
  "#666666"
)

# Assign names to match sample order
if (length(samples) <= length(sample_colors)) {
  sample_colors <- sample_colors[1:length(samples)]
  names(sample_colors) <- samples
} else {
  # If more samples than colors, extend
  names(sample_colors) <- samples[1:length(sample_colors)]
  extra_colors <- colorRampPalette(c("#1F78B4", "#33A02C", "#FB9A99"))(length(samples) - length(sample_colors))
  names(extra_colors) <- samples[(length(sample_colors)+1):length(samples)]
  sample_colors <- c(sample_colors, extra_colors)
}

# Print to verify
print("Sample colors assigned:")
print(sample_colors)

# Colors for cell types
celltype_colors <- c(
  "Tumor" = "#E41A1C",
  "Oligodendrocyte" = "#377EB8",
  "Astrocyte" = "#4DAF4A",
  "Neurons" = "#984EA3",
  "Endothelial" = "#FF7F00",
  "Myeloid/microglial" = "#FFFF33",
  "Myeloid" = "#A65628",
  "Mesenchymal" = "#F781BF",
  "Erythroid" = "#999999"
)

# Colors for neuron subtypes (from existing code)
neuron_subclass_colors <- c(
  "L2/3 IT"    = "#FEC89A",
  "L4 IT"      = "#FCD5CE",
  "L5 IT"      = "#FFDAB9",
  "L6 IT"      = "#F49595",
  "L6 IT Car3" = "#F1CEBE",
  "L5 ET"      = "#F3D17C",
  "L6 CT"      = "#FCEFB4",
  "L5/6 NP"    = "#F9ED85",
  "L6b"        = "#FFCAE9",
  "Chandelier" = "#A9E4DE",
  "Sst Chodl"  = "#83B8C6",
  "Lamp5"      = "#D0E444",
  "Lamp5 Lhx6" = "#D0EDEF",
  "Pax6"       = "#E1D3F8",
  "Vip"        = "#B0B5ED",
  "Sncg"       = "#D6DFE8",
  "Pvalb"      = "#8FC0E3",
  "Sst"        = "#7FB3D5"
)
```

---

# ALL SAMPLES ANALYSIS

## 3. Full UMAP Colored by Sample (All Cells)

```{r all_samples_umap}
# Create UMAP colored by sample
p_all_samples <- DimPlot(obj, group.by = "exp", cols = sample_colors, label = FALSE, pt.size = 0.1) +
  ggtitle("All Cells UMAP Colored by Sample") +
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5)) +
  NoAxes()

print(p_all_samples)

# Save
ggsave(filename = file.path(output_dir, "all_samples_umap.pdf"),
       plot = p_all_samples,
       device = "pdf",
       width = 12, height = 8,
       dpi = 300)

print(paste("Saved:", file.path(output_dir, "all_samples_umap.pdf")))
```

## 4. Grid of Individual Sample UMAPs (All Cells)

Create a 2-row grid where:
- Row 1: High 1/f samples
- Row 2: Low 1/f samples
- Columns: One per patient

```{r all_samples_grid}
# Get UMAP coordinates
umap_coords <- as.data.frame(Embeddings(obj, reduction = "umap"))
colnames(umap_coords) <- c("UMAP_1", "UMAP_2")
umap_coords$cell_id <- rownames(umap_coords)

# Add metadata
umap_data <- cbind(umap_coords, obj@meta.data[rownames(umap_coords), c("exp", "patient_id", "one_over_f_status", "manual_anot", "refined_annot")])

# Get UMAP limits for consistent axes
umap_xlim <- range(umap_data$UMAP_1)
umap_ylim <- range(umap_data$UMAP_2)

# Create individual plots for each sample - TWO VERSIONS: labeled and unlabeled
plot_list_labeled <- list()
plot_list_unlabeled <- list()

for (sample in samples) {
  sample_data <- umap_data[umap_data$exp == sample, ]
  n_cells <- nrow(sample_data)
  
  # LABELED version (with sample name and n)
  p_labeled <- ggplot(sample_data, aes(x = UMAP_1, y = UMAP_2, color = manual_anot)) +
    geom_point(size = 0.3, alpha = 0.8) +
    scale_color_manual(values = celltype_colors) +
    xlim(umap_xlim) + ylim(umap_ylim) +
    ggtitle(paste0(sample, "\n(n=", format(n_cells, big.mark = ","), ")")) +
    theme_void() +
    theme(plot.title = element_text(hjust = 0.5, size = 8),
          legend.position = "none")
  
  # UNLABELED version (no title)
  p_unlabeled <- ggplot(sample_data, aes(x = UMAP_1, y = UMAP_2, color = manual_anot)) +
    geom_point(size = 0.3, alpha = 0.8) +
    scale_color_manual(values = celltype_colors) +
    xlim(umap_xlim) + ylim(umap_ylim) +
    theme_void() +
    theme(legend.position = "none")
  
  plot_list_labeled[[sample]] <- p_labeled
  plot_list_unlabeled[[sample]] <- p_unlabeled
}

# Organize plots by patient (columns) and 1/f status (rows)
# Get sample order: for each patient, high1overf first then low1overf
sample_order <- umap_data %>%
  select(exp, patient_id, one_over_f_status) %>%
  distinct() %>%
  arrange(patient_id, desc(one_over_f_status)) # high before low

# Create a matrix layout
high_samples <- sample_order$exp[sample_order$one_over_f_status == "high1overf"]
low_samples <- sample_order$exp[sample_order$one_over_f_status == "low1overf"]

# --- LABELED VERSION ---
high_row_labeled <- wrap_plots(plot_list_labeled[high_samples], nrow = 1) + 
  plot_annotation(title = "High 1/f")
low_row_labeled <- wrap_plots(plot_list_labeled[low_samples], nrow = 1) + 
  plot_annotation(title = "Low 1/f")

p_grid_all_labeled <- high_row_labeled / low_row_labeled + 
  plot_annotation(title = "All Cells: Individual Sample UMAPs",
                  theme = theme(plot.title = element_text(hjust = 0.5, size = 14)))

print(p_grid_all_labeled)

ggsave(filename = file.path(output_dir, "all_samples_grid_by_patient_labeled.pdf"),
       plot = p_grid_all_labeled,
       device = "pdf",
       width = 16, height = 8,
       dpi = 300)
print(paste("Saved:", file.path(output_dir, "all_samples_grid_by_patient_labeled.pdf")))

# --- UNLABELED VERSION ---
high_row_unlabeled <- wrap_plots(plot_list_unlabeled[high_samples], nrow = 1)
low_row_unlabeled <- wrap_plots(plot_list_unlabeled[low_samples], nrow = 1)

p_grid_all_unlabeled <- high_row_unlabeled / low_row_unlabeled

print(p_grid_all_unlabeled)

ggsave(filename = file.path(output_dir, "all_samples_grid_by_patient.pdf"),
       plot = p_grid_all_unlabeled,
       device = "pdf",
       width = 16, height = 8,
       dpi = 300)
print(paste("Saved:", file.path(output_dir, "all_samples_grid_by_patient.pdf")))
```

## 5. Cell Counts CSV (All Samples)

```{r all_samples_csv}
# Create cell count table using manual_anot (broad categories): samples x cell types
cell_counts_all <- as.data.frame.matrix(table(obj@meta.data$exp, obj@meta.data$manual_anot))
cell_counts_all$sample_id <- rownames(cell_counts_all)
cell_counts_all$total <- rowSums(cell_counts_all[, -ncol(cell_counts_all)])

# Add patient_id and one_over_f_status
sample_info <- obj@meta.data %>%
  select(exp, patient_id, one_over_f_status, cortex_type) %>%
  distinct()
rownames(sample_info) <- sample_info$exp

cell_counts_all$patient_id <- sample_info[cell_counts_all$sample_id, "patient_id"]
cell_counts_all$one_over_f_status <- sample_info[cell_counts_all$sample_id, "one_over_f_status"]
cell_counts_all$cortex_type <- sample_info[cell_counts_all$sample_id, "cortex_type"]

# Reorder columns
cell_counts_all <- cell_counts_all %>%
  select(sample_id, patient_id, one_over_f_status, cortex_type, total, everything())

# Sort by patient and 1/f status
cell_counts_all <- cell_counts_all %>%
  arrange(patient_id, desc(one_over_f_status))

# Save
write.csv(cell_counts_all, 
          file.path(output_dir, "all_samples_cell_counts.csv"), 
          row.names = FALSE)

print("All Samples Cell Counts:")
print(cell_counts_all)

print(paste("Saved:", file.path(output_dir, "all_samples_cell_counts.csv")))
```

## 5b. Cell Counts CSV (Refined Annotations)

This uses `refined_annot` which combines manual annotations with MapMyCells neuron subtypes.

```{r refined_annot_csv}
# Create cell count table using refined_annot: samples x refined cell types
cell_counts_refined <- as.data.frame.matrix(table(obj@meta.data$exp, obj@meta.data$refined_annot))
cell_counts_refined$sample_id <- rownames(cell_counts_refined)
cell_counts_refined$total <- rowSums(cell_counts_refined[, -ncol(cell_counts_refined)])

# Add patient_id and one_over_f_status
cell_counts_refined$patient_id <- sample_info[cell_counts_refined$sample_id, "patient_id"]
cell_counts_refined$one_over_f_status <- sample_info[cell_counts_refined$sample_id, "one_over_f_status"]
cell_counts_refined$cortex_type <- sample_info[cell_counts_refined$sample_id, "cortex_type"]

# Reorder columns
cell_counts_refined <- cell_counts_refined %>%
  select(sample_id, patient_id, one_over_f_status, cortex_type, total, everything())

# Sort by patient and 1/f status
cell_counts_refined <- cell_counts_refined %>%
  arrange(patient_id, desc(one_over_f_status))

# Save
write.csv(cell_counts_refined, 
          file.path(output_dir, "all_samples_refined_cell_counts.csv"), 
          row.names = FALSE)

print("All Samples Refined Cell Counts (includes neuron subtypes):")
print(cell_counts_refined)

print(paste("Saved:", file.path(output_dir, "all_samples_refined_cell_counts.csv")))
```

---

# NEURONS-ONLY ANALYSIS

## 6. Subset to Neurons

```{r subset_neurons}
# Subset to neurons only
neuron_obj <- subset(obj, manual_anot == "Neurons")

print(paste("Total neurons:", ncol(neuron_obj)))
print("Neurons per sample:")
table(neuron_obj@meta.data$exp)
```

## 7. Full UMAP Colored by Sample (Neurons Only)

```{r neurons_umap}
# Create UMAP colored by sample
p_neurons_samples <- DimPlot(neuron_obj, group.by = "exp", cols = sample_colors, label = FALSE, pt.size = 0.3) +
  ggtitle("Neurons Only UMAP Colored by Sample") +
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5)) +
  NoAxes()

print(p_neurons_samples)

# Save
ggsave(filename = file.path(output_dir, "neurons_only_umap.pdf"),
       plot = p_neurons_samples,
       device = "pdf",
       width = 12, height = 8,
       dpi = 300)

print(paste("Saved:", file.path(output_dir, "neurons_only_umap.pdf")))
```

## 8. Grid of Individual Sample UMAPs (Neurons Only)

```{r neurons_grid}
# Get UMAP coordinates for neurons
neuron_umap_coords <- as.data.frame(Embeddings(neuron_obj, reduction = "umap"))
colnames(neuron_umap_coords) <- c("UMAP_1", "UMAP_2")
neuron_umap_coords$cell_id <- rownames(neuron_umap_coords)

# Add metadata
neuron_umap_data <- cbind(neuron_umap_coords, 
                          neuron_obj@meta.data[rownames(neuron_umap_coords), 
                                               c("exp", "patient_id", "one_over_f_status", "subclass_name")])

# Get UMAP limits
neuron_xlim <- range(neuron_umap_data$UMAP_1)
neuron_ylim <- range(neuron_umap_data$UMAP_2)

# Get samples that have neurons
neuron_samples <- sort(unique(neuron_umap_data$exp))

# Create individual plots for each sample - TWO VERSIONS: labeled and unlabeled
neuron_plot_list_labeled <- list()
neuron_plot_list_unlabeled <- list()

for (sample in neuron_samples) {
  sample_data <- neuron_umap_data[neuron_umap_data$exp == sample, ]
  n_cells <- nrow(sample_data)
  
  if (n_cells == 0) next
  
  # LABELED version (with sample name and n)
  p_labeled <- ggplot(sample_data, aes(x = UMAP_1, y = UMAP_2, color = subclass_name)) +
    geom_point(size = 0.5, alpha = 0.8) +
    scale_color_manual(values = neuron_subclass_colors, na.value = "darkgray") +
    xlim(neuron_xlim) + ylim(neuron_ylim) +
    ggtitle(paste0(sample, "\n(n=", format(n_cells, big.mark = ","), ")")) +
    theme_void() +
    theme(plot.title = element_text(hjust = 0.5, size = 8),
          legend.position = "none")
  
  # UNLABELED version (no title)
  p_unlabeled <- ggplot(sample_data, aes(x = UMAP_1, y = UMAP_2, color = subclass_name)) +
    geom_point(size = 0.5, alpha = 0.8) +
    scale_color_manual(values = neuron_subclass_colors, na.value = "darkgray") +
    xlim(neuron_xlim) + ylim(neuron_ylim) +
    theme_void() +
    theme(legend.position = "none")
  
  neuron_plot_list_labeled[[sample]] <- p_labeled
  neuron_plot_list_unlabeled[[sample]] <- p_unlabeled
}

# Organize by patient and 1/f status
neuron_sample_order <- neuron_umap_data %>%
  select(exp, patient_id, one_over_f_status) %>%
  distinct() %>%
  arrange(patient_id, desc(one_over_f_status))

high_neuron_samples <- neuron_sample_order$exp[neuron_sample_order$one_over_f_status == "high1overf"]
low_neuron_samples <- neuron_sample_order$exp[neuron_sample_order$one_over_f_status == "low1overf"]

# Filter to samples that have plots
high_neuron_samples <- high_neuron_samples[high_neuron_samples %in% names(neuron_plot_list_labeled)]
low_neuron_samples <- low_neuron_samples[low_neuron_samples %in% names(neuron_plot_list_labeled)]

# --- LABELED VERSION ---
if (length(high_neuron_samples) > 0) {
  high_neuron_row_labeled <- wrap_plots(neuron_plot_list_labeled[high_neuron_samples], nrow = 1) + 
    plot_annotation(title = "High 1/f")
} else {
  high_neuron_row_labeled <- plot_spacer()
}

if (length(low_neuron_samples) > 0) {
  low_neuron_row_labeled <- wrap_plots(neuron_plot_list_labeled[low_neuron_samples], nrow = 1) + 
    plot_annotation(title = "Low 1/f")
} else {
  low_neuron_row_labeled <- plot_spacer()
}

p_grid_neurons_labeled <- high_neuron_row_labeled / low_neuron_row_labeled + 
  plot_annotation(title = "Neurons Only: Individual Sample UMAPs",
                  theme = theme(plot.title = element_text(hjust = 0.5, size = 14)))

print(p_grid_neurons_labeled)

ggsave(filename = file.path(output_dir, "neurons_only_grid_by_patient_labeled.pdf"),
       plot = p_grid_neurons_labeled,
       device = "pdf",
       width = 16, height = 8,
       dpi = 300)
print(paste("Saved:", file.path(output_dir, "neurons_only_grid_by_patient_labeled.pdf")))

# --- UNLABELED VERSION ---
if (length(high_neuron_samples) > 0) {
  high_neuron_row_unlabeled <- wrap_plots(neuron_plot_list_unlabeled[high_neuron_samples], nrow = 1)
} else {
  high_neuron_row_unlabeled <- plot_spacer()
}

if (length(low_neuron_samples) > 0) {
  low_neuron_row_unlabeled <- wrap_plots(neuron_plot_list_unlabeled[low_neuron_samples], nrow = 1)
} else {
  low_neuron_row_unlabeled <- plot_spacer()
}

p_grid_neurons_unlabeled <- high_neuron_row_unlabeled / low_neuron_row_unlabeled

print(p_grid_neurons_unlabeled)

ggsave(filename = file.path(output_dir, "neurons_only_grid_by_patient.pdf"),
       plot = p_grid_neurons_unlabeled,
       device = "pdf",
       width = 16, height = 8,
       dpi = 300)
print(paste("Saved:", file.path(output_dir, "neurons_only_grid_by_patient.pdf")))
```

## 9. Cell Counts CSV (Neurons Only - by Subclass)

```{r neurons_csv}
# Create cell count table: samples x neuron subtypes (subclass_name)
neuron_counts <- as.data.frame.matrix(table(neuron_obj@meta.data$exp, 
                                            neuron_obj@meta.data$subclass_name, 
                                            useNA = "ifany"))

# Rename NA column if exists
if ("NA" %in% colnames(neuron_counts) || any(is.na(colnames(neuron_counts)))) {
  na_col <- which(is.na(colnames(neuron_counts)) | colnames(neuron_counts) == "NA")
  if (length(na_col) > 0) {
    colnames(neuron_counts)[na_col] <- "Unassigned"
  }
}

neuron_counts$sample_id <- rownames(neuron_counts)
neuron_counts$total <- rowSums(neuron_counts[, -ncol(neuron_counts)])

# Add sample info
neuron_sample_info <- neuron_obj@meta.data %>%
  select(exp, patient_id, one_over_f_status, cortex_type) %>%
  distinct()
rownames(neuron_sample_info) <- neuron_sample_info$exp

neuron_counts$patient_id <- neuron_sample_info[neuron_counts$sample_id, "patient_id"]
neuron_counts$one_over_f_status <- neuron_sample_info[neuron_counts$sample_id, "one_over_f_status"]
neuron_counts$cortex_type <- neuron_sample_info[neuron_counts$sample_id, "cortex_type"]

# Reorder columns
neuron_counts <- neuron_counts %>%
  select(sample_id, patient_id, one_over_f_status, cortex_type, total, everything())

# Sort by patient and 1/f status
neuron_counts <- neuron_counts %>%
  arrange(patient_id, desc(one_over_f_status))

# Save
write.csv(neuron_counts, 
          file.path(output_dir, "neurons_only_cell_counts.csv"), 
          row.names = FALSE)

print("Neurons Only Cell Counts (by subclass):")
print(neuron_counts)

print(paste("Saved:", file.path(output_dir, "neurons_only_cell_counts.csv")))
```

## 10. Additional: Cell Counts by Class (Glutamatergic vs GABAergic)

```{r neurons_class_csv}
# Create cell count table: samples x class_name
neuron_class_counts <- as.data.frame.matrix(table(neuron_obj@meta.data$exp, 
                                                   neuron_obj@meta.data$class_name, 
                                                   useNA = "ifany"))

# Rename NA column if exists
if (any(is.na(colnames(neuron_class_counts)))) {
  na_col <- which(is.na(colnames(neuron_class_counts)))
  colnames(neuron_class_counts)[na_col] <- "Unassigned"
}

neuron_class_counts$sample_id <- rownames(neuron_class_counts)
neuron_class_counts$total <- rowSums(neuron_class_counts[, -ncol(neuron_class_counts)])

# Add sample info
neuron_class_counts$patient_id <- neuron_sample_info[neuron_class_counts$sample_id, "patient_id"]
neuron_class_counts$one_over_f_status <- neuron_sample_info[neuron_class_counts$sample_id, "one_over_f_status"]
neuron_class_counts$cortex_type <- neuron_sample_info[neuron_class_counts$sample_id, "cortex_type"]

# Reorder columns
neuron_class_counts <- neuron_class_counts %>%
  select(sample_id, patient_id, one_over_f_status, cortex_type, total, everything())

# Sort
neuron_class_counts <- neuron_class_counts %>%
  arrange(patient_id, desc(one_over_f_status))

# Save
write.csv(neuron_class_counts, 
          file.path(output_dir, "neurons_only_class_counts.csv"), 
          row.names = FALSE)

print("Neurons Only Cell Counts (by class - Glutamatergic vs GABAergic):")
print(neuron_class_counts)

print(paste("Saved:", file.path(output_dir, "neurons_only_class_counts.csv")))
```

## Summary

```{r summary}
print("=== OUTPUT SUMMARY ===")
print("")
print("PDFs created:")
print(paste("  1.", file.path(output_dir, "all_samples_umap.pdf")))
print(paste("  2.", file.path(output_dir, "all_samples_grid_by_patient.pdf"), "(no labels)"))
print(paste("  3.", file.path(output_dir, "all_samples_grid_by_patient_labeled.pdf"), "(with labels)"))
print(paste("  4.", file.path(output_dir, "neurons_only_umap.pdf")))
print(paste("  5.", file.path(output_dir, "neurons_only_grid_by_patient.pdf"), "(no labels)"))
print(paste("  6.", file.path(output_dir, "neurons_only_grid_by_patient_labeled.pdf"), "(with labels)"))
print("")
print("CSVs created:")
print(paste("  1.", file.path(output_dir, "all_samples_cell_counts.csv"), "(broad cell types)"))
print(paste("  2.", file.path(output_dir, "all_samples_refined_cell_counts.csv"), "(with neuron subtypes)"))
print(paste("  3.", file.path(output_dir, "neurons_only_cell_counts.csv"), "(neuron subclasses)"))
print(paste("  4.", file.path(output_dir, "neurons_only_class_counts.csv"), "(Glut vs GABA)"))
```

## Session Info

```{r session_info}
sessionInfo()
```
