---
title: "S3: Epilepsy Sample-Level UMAP Analysis"
output: html_notebook
---

This script generates sample-level UMAPs and cell count reports for the epilepsy cohort:
- Full UMAP colored by sample (all cells + neurons only)
- Grid of individual sample UMAPs organized by 1/f status
- CSV reports of cell counts by sample and cell type

## Setup

```{r setup}
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
library(tidyr)
library(RColorBrewer)
```

## 1. Load Epilepsy Object

```{r load_object}
# Load the epilepsy object with MapMyCells annotations
obj <- readRDS("D:/Youssef Sibih/Seurat_Objects/R1_123125/epilepsy_4samples_combined_sct_MapCells.rds")

# Create output directory if needed
output_dir <- "d:/AbrahamDada/oneoverf/oneoverf_R1_10326/plots"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# Quick verification
print(paste("Total cells:", ncol(obj)))
print("Metadata columns:")
print(colnames(obj@meta.data))
```

## 2. Identify Sample Field and Verify Metadata

```{r identify_samples}
# Check common sample identifier fields
# Prioritize 'exp' as it contains actual sample names (e.g., SF694D_4)
# 'orig.ident' is often just the project name (e.g., "1overF_snRNA")
if ("exp" %in% colnames(obj@meta.data)) {
  sample_field <- "exp"
  print("Using 'exp' as sample identifier")
} else if ("orig.ident" %in% colnames(obj@meta.data)) {
  sample_field <- "orig.ident"
  print("Using 'orig.ident' as sample identifier")
} else {
  # List potential sample fields
  print("Available metadata columns:")
  print(colnames(obj@meta.data))
  stop("Could not identify sample field. Please set sample_field manually.")
}

# Get unique samples
samples <- sort(unique(obj@meta.data[[sample_field]]))
print(paste("Samples found:", paste(samples, collapse = ", ")))
print(paste("Number of samples:", length(samples)))

# Check one_over_f_status
if ("one_over_f_status" %in% colnames(obj@meta.data)) {
  print("one_over_f_status distribution:")
  print(table(obj@meta.data$one_over_f_status))
} else {
  print("Warning: one_over_f_status not found in metadata")
}

# Check class_name (MapMyCells)
if ("class_name" %in% colnames(obj@meta.data)) {
  print("class_name distribution:")
  print(table(obj@meta.data$class_name, useNA = "ifany"))
}
```

## 3. Define Color Palettes

```{r color_palettes}
# Create a distinct color palette for samples
# Using first N colors from a highly distinguishable palette
base_colors <- c(
  "#E41A1C",
"#377EB8",
  "#4DAF4A",
  "#984EA3",
  "#FF7F00",
  "#A65628",
  "#F781BF",
  "#1B9E77"
)

# Assign colors to samples
n_samples <- length(samples)
if (n_samples <= length(base_colors)) {
  sample_colors <- base_colors[1:n_samples]
} else {
  sample_colors <- colorRampPalette(base_colors)(n_samples)
}
names(sample_colors) <- samples

print("Sample colors assigned:")
print(sample_colors)

# Colors for MapMyCells class types
class_colors <- c(
  "Neuronal: Glutamatergic" = "#FF7F00",
  "Neuronal: GABAergic" = "#377EB8",
  "Non-neuronal and Non-neural" = "#999999"
)

# Colors for neuron subtypes (from existing code)
neuron_subclass_colors <- c(
  "L2/3 IT"    = "#FEC89A",
  "L4 IT"      = "#FCD5CE",
  "L5 IT"      = "#FFDAB9",
  "L6 IT"      = "#F49595",
  "L6 IT Car3" = "#F1CEBE",
  "L5 ET"      = "#F3D17C",
  "L6 CT"      = "#FCEFB4",
  "L5/6 NP"    = "#F9ED85",
  "L6b"        = "#FFCAE9",
  "Chandelier" = "#A9E4DE",
  "Sst Chodl"  = "#83B8C6",
  "Lamp5"      = "#D0E444",
  "Lamp5 Lhx6" = "#D0EDEF",
  "Pax6"       = "#E1D3F8",
  "Vip"        = "#B0B5ED",
  "Sncg"       = "#D6DFE8",
  "Pvalb"      = "#8FC0E3",
  "Sst"        = "#7FB3D5"
)
```

---

# ALL CELLS ANALYSIS

## 4. Full UMAP Colored by Sample (All Cells)

```{r all_samples_umap}
# Create UMAP colored by sample
p_all_samples <- DimPlot(obj, group.by = sample_field, cols = sample_colors, label = FALSE, pt.size = 0.1) +
  ggtitle("Epilepsy: All Cells UMAP Colored by Sample") +
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5)) +
  NoAxes()

print(p_all_samples)

# Save
ggsave(filename = file.path(output_dir, "epilepsy_all_samples_umap.pdf"),
       plot = p_all_samples,
       device = "pdf",
       width = 10, height = 8,
       dpi = 300)

print(paste("Saved:", file.path(output_dir, "epilepsy_all_samples_umap.pdf")))
```

## 5. Grid of Individual Sample UMAPs (All Cells)

```{r all_samples_grid}
# Get UMAP coordinates
umap_coords <- as.data.frame(Embeddings(obj, reduction = "umap"))
colnames(umap_coords) <- c("UMAP_1", "UMAP_2")
umap_coords$cell_id <- rownames(umap_coords)

# Add metadata
metadata_cols <- c(sample_field, "one_over_f_status", "class_name")
metadata_cols <- metadata_cols[metadata_cols %in% colnames(obj@meta.data)]
umap_data <- cbind(umap_coords, obj@meta.data[rownames(umap_coords), metadata_cols, drop = FALSE])

# Rename sample column for consistency
colnames(umap_data)[colnames(umap_data) == sample_field] <- "sample_id"

# Get UMAP limits for consistent axes
umap_xlim <- range(umap_data$UMAP_1)
umap_ylim <- range(umap_data$UMAP_2)

# Create individual plots for each sample - TWO VERSIONS: labeled and unlabeled
plot_list_labeled <- list()
plot_list_unlabeled <- list()

for (sample in samples) {
  sample_data <- umap_data[umap_data$sample_id == sample, ]
  n_cells <- nrow(sample_data)
  
  # LABELED version (with sample name and n)
  p_labeled <- ggplot(sample_data, aes(x = UMAP_1, y = UMAP_2, color = class_name)) +
    geom_point(size = 0.3, alpha = 0.8) +
    scale_color_manual(values = class_colors, na.value = "gray50") +
    xlim(umap_xlim) + ylim(umap_ylim) +
    ggtitle(paste0(sample, "\n(n=", format(n_cells, big.mark = ","), ")")) +
    theme_void() +
    theme(plot.title = element_text(hjust = 0.5, size = 8),
          legend.position = "none")
  
  # UNLABELED version (no title)
  p_unlabeled <- ggplot(sample_data, aes(x = UMAP_1, y = UMAP_2, color = class_name)) +
    geom_point(size = 0.3, alpha = 0.8) +
    scale_color_manual(values = class_colors, na.value = "gray50") +
    xlim(umap_xlim) + ylim(umap_ylim) +
    theme_void() +
    theme(legend.position = "none")
  
  plot_list_labeled[[sample]] <- p_labeled
  plot_list_unlabeled[[sample]] <- p_unlabeled
}

# Organize plots by 1/f status (rows)
if ("one_over_f_status" %in% colnames(umap_data)) {
  sample_order <- umap_data %>%
    select(sample_id, one_over_f_status) %>%
    distinct() %>%
    arrange(desc(one_over_f_status), sample_id) # high before low
  
  high_samples <- sample_order$sample_id[sample_order$one_over_f_status == "high1overf"]
  low_samples <- sample_order$sample_id[sample_order$one_over_f_status == "low1overf"]
  
  # --- LABELED VERSION ---
  if (length(high_samples) > 0) {
    high_row_labeled <- wrap_plots(plot_list_labeled[high_samples], nrow = 1) + 
      plot_annotation(title = "High 1/f")
  } else {
    high_row_labeled <- plot_spacer()
  }
  
  if (length(low_samples) > 0) {
    low_row_labeled <- wrap_plots(plot_list_labeled[low_samples], nrow = 1) + 
      plot_annotation(title = "Low 1/f")
  } else {
    low_row_labeled <- plot_spacer()
  }
  
  p_grid_all_labeled <- high_row_labeled / low_row_labeled + 
    plot_annotation(title = "Epilepsy All Cells: Individual Sample UMAPs",
                    theme = theme(plot.title = element_text(hjust = 0.5, size = 14)))
  
  # --- UNLABELED VERSION ---
  if (length(high_samples) > 0) {
    high_row_unlabeled <- wrap_plots(plot_list_unlabeled[high_samples], nrow = 1)
  } else {
    high_row_unlabeled <- plot_spacer()
  }
  
  if (length(low_samples) > 0) {
    low_row_unlabeled <- wrap_plots(plot_list_unlabeled[low_samples], nrow = 1)
  } else {
    low_row_unlabeled <- plot_spacer()
  }
  
  p_grid_all_unlabeled <- high_row_unlabeled / low_row_unlabeled
  
} else {
  # No 1/f status - just arrange all samples in one row
  p_grid_all_labeled <- wrap_plots(plot_list_labeled, nrow = 1) + 
    plot_annotation(title = "Epilepsy All Cells: Individual Sample UMAPs",
                    theme = theme(plot.title = element_text(hjust = 0.5, size = 14)))
  p_grid_all_unlabeled <- wrap_plots(plot_list_unlabeled, nrow = 1)
}

print(p_grid_all_labeled)

# Save labeled version
ggsave(filename = file.path(output_dir, "epilepsy_all_samples_grid_by_patient_labeled.pdf"),
       plot = p_grid_all_labeled,
       device = "pdf",
       width = 12, height = 8,
       dpi = 300)
print(paste("Saved:", file.path(output_dir, "epilepsy_all_samples_grid_by_patient_labeled.pdf")))

# Save unlabeled version
ggsave(filename = file.path(output_dir, "epilepsy_all_samples_grid_by_patient.pdf"),
       plot = p_grid_all_unlabeled,
       device = "pdf",
       width = 12, height = 8,
       dpi = 300)
print(paste("Saved:", file.path(output_dir, "epilepsy_all_samples_grid_by_patient.pdf")))
```

## 6. Cell Counts CSV (All Samples)

```{r all_samples_csv}
# Create cell count table using class_name: samples x cell types
cell_counts_all <- as.data.frame.matrix(table(obj@meta.data[[sample_field]], obj@meta.data$class_name))
cell_counts_all$sample_id <- rownames(cell_counts_all)
cell_counts_all$total <- rowSums(cell_counts_all[, -ncol(cell_counts_all)])

# Add one_over_f_status if available
if ("one_over_f_status" %in% colnames(obj@meta.data)) {
  sample_info <- obj@meta.data %>%
    select(all_of(c(sample_field, "one_over_f_status"))) %>%
    distinct()
  colnames(sample_info)[1] <- "sample_id"
  rownames(sample_info) <- sample_info$sample_id
  
  cell_counts_all$one_over_f_status <- sample_info[cell_counts_all$sample_id, "one_over_f_status"]
  
  # Reorder columns
  cell_counts_all <- cell_counts_all %>%
    select(sample_id, one_over_f_status, total, everything())
  
  # Sort by 1/f status
  cell_counts_all <- cell_counts_all %>%
    arrange(desc(one_over_f_status), sample_id)
} else {
  cell_counts_all <- cell_counts_all %>%
    select(sample_id, total, everything())
}

# Save
write.csv(cell_counts_all, 
          file.path(output_dir, "epilepsy_all_samples_cell_counts.csv"), 
          row.names = FALSE)

print("Epilepsy All Samples Cell Counts:")
print(cell_counts_all)

print(paste("Saved:", file.path(output_dir, "epilepsy_all_samples_cell_counts.csv")))
```

---

# NEURONS-ONLY ANALYSIS

## 7. Subset to Neurons

```{r subset_neurons}
# Subset to neurons only using MapMyCells class_name
neuron_obj <- subset(obj, subset = class_name %in% c("Neuronal: Glutamatergic", "Neuronal: GABAergic"))

print(paste("Total neurons:", ncol(neuron_obj)))
print("Neurons per sample:")
print(table(neuron_obj@meta.data[[sample_field]]))

print("Neuron subclass distribution:")
print(table(neuron_obj@meta.data$subclass_name, useNA = "ifany"))
```

## 8. Full UMAP Colored by Sample (Neurons Only)

```{r neurons_umap}
# Create UMAP colored by sample
p_neurons_samples <- DimPlot(neuron_obj, group.by = sample_field, cols = sample_colors, label = FALSE, pt.size = 0.3) +
  ggtitle("Epilepsy: Neurons Only UMAP Colored by Sample") +
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5)) +
  NoAxes()

print(p_neurons_samples)

# Save
ggsave(filename = file.path(output_dir, "epilepsy_neurons_only_umap.pdf"),
       plot = p_neurons_samples,
       device = "pdf",
       width = 10, height = 8,
       dpi = 300)

print(paste("Saved:", file.path(output_dir, "epilepsy_neurons_only_umap.pdf")))
```

## 9. Grid of Individual Sample UMAPs (Neurons Only)

```{r neurons_grid}
# Get UMAP coordinates for neurons
neuron_umap_coords <- as.data.frame(Embeddings(neuron_obj, reduction = "umap"))
colnames(neuron_umap_coords) <- c("UMAP_1", "UMAP_2")
neuron_umap_coords$cell_id <- rownames(neuron_umap_coords)

# Add metadata
neuron_metadata_cols <- c(sample_field, "one_over_f_status", "subclass_name")
neuron_metadata_cols <- neuron_metadata_cols[neuron_metadata_cols %in% colnames(neuron_obj@meta.data)]
neuron_umap_data <- cbind(neuron_umap_coords, 
                          neuron_obj@meta.data[rownames(neuron_umap_coords), neuron_metadata_cols, drop = FALSE])

# Rename sample column
colnames(neuron_umap_data)[colnames(neuron_umap_data) == sample_field] <- "sample_id"

# Get UMAP limits
neuron_xlim <- range(neuron_umap_data$UMAP_1)
neuron_ylim <- range(neuron_umap_data$UMAP_2)

# Get samples that have neurons
neuron_samples <- sort(unique(neuron_umap_data$sample_id))

# Create individual plots for each sample - TWO VERSIONS: labeled and unlabeled
neuron_plot_list_labeled <- list()
neuron_plot_list_unlabeled <- list()

for (sample in neuron_samples) {
  sample_data <- neuron_umap_data[neuron_umap_data$sample_id == sample, ]
  n_cells <- nrow(sample_data)
  
  if (n_cells == 0) next
  
  # LABELED version (with sample name and n)
  p_labeled <- ggplot(sample_data, aes(x = UMAP_1, y = UMAP_2, color = subclass_name)) +
    geom_point(size = 0.5, alpha = 0.8) +
    scale_color_manual(values = neuron_subclass_colors, na.value = "darkgray") +
    xlim(neuron_xlim) + ylim(neuron_ylim) +
    ggtitle(paste0(sample, "\n(n=", format(n_cells, big.mark = ","), ")")) +
    theme_void() +
    theme(plot.title = element_text(hjust = 0.5, size = 8),
          legend.position = "none")
  
  # UNLABELED version (no title)
  p_unlabeled <- ggplot(sample_data, aes(x = UMAP_1, y = UMAP_2, color = subclass_name)) +
    geom_point(size = 0.5, alpha = 0.8) +
    scale_color_manual(values = neuron_subclass_colors, na.value = "darkgray") +
    xlim(neuron_xlim) + ylim(neuron_ylim) +
    theme_void() +
    theme(legend.position = "none")
  
  neuron_plot_list_labeled[[sample]] <- p_labeled
  neuron_plot_list_unlabeled[[sample]] <- p_unlabeled
}

# Organize by 1/f status
if ("one_over_f_status" %in% colnames(neuron_umap_data)) {
  neuron_sample_order <- neuron_umap_data %>%
    select(sample_id, one_over_f_status) %>%
    distinct() %>%
    arrange(desc(one_over_f_status), sample_id)
  
  high_neuron_samples <- neuron_sample_order$sample_id[neuron_sample_order$one_over_f_status == "high1overf"]
  low_neuron_samples <- neuron_sample_order$sample_id[neuron_sample_order$one_over_f_status == "low1overf"]
  
  # Filter to samples that have plots
  high_neuron_samples <- high_neuron_samples[high_neuron_samples %in% names(neuron_plot_list_labeled)]
  low_neuron_samples <- low_neuron_samples[low_neuron_samples %in% names(neuron_plot_list_labeled)]
  
  # --- LABELED VERSION ---
  if (length(high_neuron_samples) > 0) {
    high_neuron_row_labeled <- wrap_plots(neuron_plot_list_labeled[high_neuron_samples], nrow = 1) + 
      plot_annotation(title = "High 1/f")
  } else {
    high_neuron_row_labeled <- plot_spacer()
  }
  
  if (length(low_neuron_samples) > 0) {
    low_neuron_row_labeled <- wrap_plots(neuron_plot_list_labeled[low_neuron_samples], nrow = 1) + 
      plot_annotation(title = "Low 1/f")
  } else {
    low_neuron_row_labeled <- plot_spacer()
  }
  
  p_grid_neurons_labeled <- high_neuron_row_labeled / low_neuron_row_labeled + 
    plot_annotation(title = "Epilepsy Neurons Only: Individual Sample UMAPs",
                    theme = theme(plot.title = element_text(hjust = 0.5, size = 14)))
  
  # --- UNLABELED VERSION ---
  if (length(high_neuron_samples) > 0) {
    high_neuron_row_unlabeled <- wrap_plots(neuron_plot_list_unlabeled[high_neuron_samples], nrow = 1)
  } else {
    high_neuron_row_unlabeled <- plot_spacer()
  }
  
  if (length(low_neuron_samples) > 0) {
    low_neuron_row_unlabeled <- wrap_plots(neuron_plot_list_unlabeled[low_neuron_samples], nrow = 1)
  } else {
    low_neuron_row_unlabeled <- plot_spacer()
  }
  
  p_grid_neurons_unlabeled <- high_neuron_row_unlabeled / low_neuron_row_unlabeled
  
} else {
  # No 1/f status - arrange all in one row
  p_grid_neurons_labeled <- wrap_plots(neuron_plot_list_labeled, nrow = 1) + 
    plot_annotation(title = "Epilepsy Neurons Only: Individual Sample UMAPs",
                    theme = theme(plot.title = element_text(hjust = 0.5, size = 14)))
  p_grid_neurons_unlabeled <- wrap_plots(neuron_plot_list_unlabeled, nrow = 1)
}

print(p_grid_neurons_labeled)

# Save labeled version
ggsave(filename = file.path(output_dir, "epilepsy_neurons_only_grid_by_patient_labeled.pdf"),
       plot = p_grid_neurons_labeled,
       device = "pdf",
       width = 12, height = 8,
       dpi = 300)
print(paste("Saved:", file.path(output_dir, "epilepsy_neurons_only_grid_by_patient_labeled.pdf")))

# Save unlabeled version
ggsave(filename = file.path(output_dir, "epilepsy_neurons_only_grid_by_patient.pdf"),
       plot = p_grid_neurons_unlabeled,
       device = "pdf",
       width = 12, height = 8,
       dpi = 300)
print(paste("Saved:", file.path(output_dir, "epilepsy_neurons_only_grid_by_patient.pdf")))
```

## 10. Cell Counts CSV (Neurons Only - by Subclass)

```{r neurons_csv}
# Create cell count table: samples x neuron subtypes (subclass_name)
neuron_counts <- as.data.frame.matrix(table(neuron_obj@meta.data[[sample_field]], 
                                            neuron_obj@meta.data$subclass_name, 
                                            useNA = "ifany"))

# Rename NA column if exists
if (any(is.na(colnames(neuron_counts)))) {
  na_col <- which(is.na(colnames(neuron_counts)))
  colnames(neuron_counts)[na_col] <- "Unassigned"
}

neuron_counts$sample_id <- rownames(neuron_counts)
neuron_counts$total <- rowSums(neuron_counts[, -ncol(neuron_counts)])

# Add one_over_f_status if available
if ("one_over_f_status" %in% colnames(neuron_obj@meta.data)) {
  neuron_sample_info <- neuron_obj@meta.data %>%
    select(all_of(c(sample_field, "one_over_f_status"))) %>%
    distinct()
  colnames(neuron_sample_info)[1] <- "sample_id"
  rownames(neuron_sample_info) <- neuron_sample_info$sample_id
  
  neuron_counts$one_over_f_status <- neuron_sample_info[neuron_counts$sample_id, "one_over_f_status"]
  
  # Reorder columns
  neuron_counts <- neuron_counts %>%
    select(sample_id, one_over_f_status, total, everything())
  
  # Sort
  neuron_counts <- neuron_counts %>%
    arrange(desc(one_over_f_status), sample_id)
} else {
  neuron_counts <- neuron_counts %>%
    select(sample_id, total, everything())
}

# Save
write.csv(neuron_counts, 
          file.path(output_dir, "epilepsy_neurons_only_cell_counts.csv"), 
          row.names = FALSE)

print("Epilepsy Neurons Only Cell Counts (by subclass):")
print(neuron_counts)

print(paste("Saved:", file.path(output_dir, "epilepsy_neurons_only_cell_counts.csv")))
```

## 11. Cell Counts by Class (Glutamatergic vs GABAergic)

```{r neurons_class_csv}
# Create cell count table: samples x class_name
neuron_class_counts <- as.data.frame.matrix(table(neuron_obj@meta.data[[sample_field]], 
                                                   neuron_obj@meta.data$class_name, 
                                                   useNA = "ifany"))

# Rename NA column if exists
if (any(is.na(colnames(neuron_class_counts)))) {
  na_col <- which(is.na(colnames(neuron_class_counts)))
  colnames(neuron_class_counts)[na_col] <- "Unassigned"
}

neuron_class_counts$sample_id <- rownames(neuron_class_counts)
neuron_class_counts$total <- rowSums(neuron_class_counts[, -ncol(neuron_class_counts)])

# Add one_over_f_status if available
if ("one_over_f_status" %in% colnames(neuron_obj@meta.data)) {
  neuron_class_counts$one_over_f_status <- neuron_sample_info[neuron_class_counts$sample_id, "one_over_f_status"]
  
  # Reorder columns
  neuron_class_counts <- neuron_class_counts %>%
    select(sample_id, one_over_f_status, total, everything())
  
  # Sort
  neuron_class_counts <- neuron_class_counts %>%
    arrange(desc(one_over_f_status), sample_id)
} else {
  neuron_class_counts <- neuron_class_counts %>%
    select(sample_id, total, everything())
}

# Save
write.csv(neuron_class_counts, 
          file.path(output_dir, "epilepsy_neurons_only_class_counts.csv"), 
          row.names = FALSE)

print("Epilepsy Neurons Only Cell Counts (by class - Glutamatergic vs GABAergic):")
print(neuron_class_counts)

print(paste("Saved:", file.path(output_dir, "epilepsy_neurons_only_class_counts.csv")))
```

## Summary

```{r summary}
print("=== OUTPUT SUMMARY ===")
print("")
print("PDFs created:")
print(paste("  1.", file.path(output_dir, "epilepsy_all_samples_umap.pdf")))
print(paste("  2.", file.path(output_dir, "epilepsy_all_samples_grid_by_patient.pdf"), "(no labels)"))
print(paste("  3.", file.path(output_dir, "epilepsy_all_samples_grid_by_patient_labeled.pdf"), "(with labels)"))
print(paste("  4.", file.path(output_dir, "epilepsy_neurons_only_umap.pdf")))
print(paste("  5.", file.path(output_dir, "epilepsy_neurons_only_grid_by_patient.pdf"), "(no labels)"))
print(paste("  6.", file.path(output_dir, "epilepsy_neurons_only_grid_by_patient_labeled.pdf"), "(with labels)"))
print("")
print("CSVs created:")
print(paste("  1.", file.path(output_dir, "epilepsy_all_samples_cell_counts.csv"), "(by MapMyCells class)"))
print(paste("  2.", file.path(output_dir, "epilepsy_neurons_only_cell_counts.csv"), "(neuron subclasses)"))
print(paste("  3.", file.path(output_dir, "epilepsy_neurons_only_class_counts.csv"), "(Glut vs GABA)"))
```

## Session Info

```{r session_info}
sessionInfo()
```
