---
title: "S4: Sample-Level PCA of Transcriptomes"
output: html_notebook
---

This script performs sample-level PCA on pseudobulk transcriptomes to assess whether gene expression profiles segregate by:
- 1/f exponent (flat vs steep)
- Cortex type (Normal Cortex, GBM, Oligodendroglioma, Astrocytoma)

## Setup

```{r setup}
library(dplyr)
library(Seurat)
library(ggplot2)
library(ggrepel)
library(DESeq2)
```

## 1. Load Annotated Object

```{r load_object}
obj <- readRDS("D:/Youssef Sibih/Seurat_Objects/oneoverf_combined_corrected_annotations.rds")

# Create output directory if needed
output_dir <- "plots"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

print(paste("Total cells:", ncol(obj)))
print("Samples (exp):")
print(unique(obj@meta.data$exp))
```

## 2. Pseudobulk Aggregation

Aggregate raw counts per sample (one expression profile per sample).

```{r pseudobulk}
# Sum counts per gene per sample; returns matrix genes x samples
pb_counts <- Seurat::AggregateExpression(
  obj,
  group.by = "exp",
  assays = "RNA",
  slot = "counts",
  return.seurat = FALSE
)[["RNA"]]

# Ensure we have counts (not SCT)
# If RNA counts are empty, try "SCT" with slot "counts" as fallback
if (length(pb_counts) == 0 || all(pb_counts == 0)) {
  pb_counts <- Seurat::AggregateExpression(
    obj,
    group.by = "exp",
    assays = "SCT",
    slot = "counts",
    return.seurat = FALSE
  )[["SCT"]]
}

print("Pseudobulk matrix dimensions (genes x samples):")
print(dim(pb_counts))
print("Sample columns:")
print(colnames(pb_counts))
```

## 3. Variance-Stabilizing Transform (DESeq2)

Normalize pseudobulk counts with VST for PCA.

```{r vst}
# Sample-level metadata for DESeq2 (required for colData)
sample_ids <- colnames(pb_counts)
sample_meta <- data.frame(
  sample = sample_ids,
  row.names = sample_ids
)

# Build DESeqDataSet with intercept-only design (for normalization only)
dds <- DESeqDataSetFromMatrix(
  countData = pb_counts,
  colData = sample_meta,
  design = ~ 1
)

# Filter low-count genes to avoid VST issues
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep, ]

# Variance-stabilizing transformation
vst_mat <- assay(varianceStabilizingTransformation(dds))

print("VST matrix dimensions (genes x samples):")
print(dim(vst_mat))
```

## 4. Run PCA

```{r pca}
# PCA: samples as rows, genes as columns; center and scale
pca_res <- prcomp(t(vst_mat), center = TRUE, scale. = TRUE)

# Variance explained by first 5 PCs
var_explained <- summary(pca_res)$importance[2, ]
print("Variance explained (first 5 PCs):")
print(round(var_explained[1:5], 4))
```

## 5. Sample-Level Metadata for Plotting

```{r sample_metadata}
# One row per sample: exp, one_over_f_status, cortex_type
meta_cell <- obj@meta.data
sample_meta_plot <- meta_cell %>%
  distinct(exp, .keep_all = TRUE) %>%
  select(exp, one_over_f_status, cortex_type) %>%
  mutate(
    oneoverf_label = case_when(
      one_over_f_status == "high1overf" ~ "Steep (high 1/f)",
      one_over_f_status == "low1overf"  ~ "Flat (low 1/f)",
      TRUE ~ as.character(one_over_f_status)
    )
  )

# Align with PCA rownames (sample order)
# AggregateExpression converts underscores to hyphens; revert to match metadata
sample_order <- gsub("^RNA\\.|^SCT\\.", "", rownames(pca_res$x))
sample_order <- gsub("-", "_", sample_order)
rownames(pca_res$x) <- sample_order
sample_meta_plot <- sample_meta_plot[match(sample_order, sample_meta_plot$exp), ]
sample_meta_plot$sample <- sample_meta_plot$exp

# Combine PCA coordinates with metadata
pca_df <- as.data.frame(pca_res$x[, 1:2])
pca_df$sample <- rownames(pca_df)
pca_df <- left_join(pca_df, sample_meta_plot, by = "sample")

# Percent variance for axis labels
pct1 <- round(100 * var_explained["PC1"], 1)
pct2 <- round(100 * var_explained["PC2"], 1)
```

## 6. Plot 1: PCA Colored by 1/f Status (Flat vs Steep)

```{r plot_oneoverf}
# Base plot (shared layers)
p1_base <- ggplot(pca_df, aes(x = PC1, y = PC2, color = oneoverf_label)) +
  geom_point(size = 4) +
  scale_color_manual(values = c("Flat (low 1/f)" = "#19bdc1", "Steep (high 1/f)" = "#f5746f")) +
  labs(
    x = "PC1",
    y = "PC2",
    color = "1/f status",
    title = "Sample-level PCA: transcriptomes by 1/f exponent"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "right")

# With sample labels
p1_labeled <- p1_base +
  geom_text_repel(aes(label = sample), size = 3.5, max.overlaps = 20)

# Without sample labels
p1_nolabel <- p1_base

print(p1_labeled)
print(p1_nolabel)
```

## 7. Plot 2: PCA Colored by Cortex Type

```{r plot_cortex_type}
# Friendly cortex type labels
pca_df <- pca_df %>%
  mutate(cortex_label = case_when(
    cortex_type == "Normal Cortex"     ~ "Normal Cortex",
    cortex_type == "glioblastoma"      ~ "GBM",
    cortex_type == "oligodendroglioma"  ~ "Oligodendroglioma",
    cortex_type == "astrocytoma"        ~ "Astrocytoma",
    TRUE ~ as.character(cortex_type)
  ))

p2_base <- ggplot(pca_df, aes(x = PC1, y = PC2, color = cortex_label)) +
  geom_point(size = 4) +
  scale_color_manual(
    values = c(
      "Normal Cortex"     = "#658ec7",
      "GBM"               = "#01a99d",
      "Oligodendroglioma" = "#000000",
      "Astrocytoma"       = "#e91f8c"
    )
  ) +
  labs(
    x = "PC1",
    y = "PC2",
    color = "Cortex type",
    title = "Sample-level PCA: transcriptomes by cortex type"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "right")

# With sample labels
p2_labeled <- p2_base +
  geom_text_repel(aes(label = sample), size = 3.5, max.overlaps = 20)

# Without sample labels
p2_nolabel <- p2_base

print(p2_labeled)
print(p2_nolabel)
```

## 8. Save Plots to PDF

```{r save_plots}
# With labels
ggsave(file.path(output_dir, "S4_PCA_by_oneoverf_status_labeled.pdf"), p1_labeled, width = 7, height = 5)
ggsave(file.path(output_dir, "S4_PCA_by_cortex_type_labeled.pdf"), p2_labeled, width = 7, height = 5)

# Without labels
ggsave(file.path(output_dir, "S4_PCA_by_oneoverf_status_nolabel.pdf"), p1_nolabel, width = 7, height = 5)
ggsave(file.path(output_dir, "S4_PCA_by_cortex_type_nolabel.pdf"), p2_nolabel, width = 7, height = 5)

print(paste("Plots saved to", output_dir))
```

## Session Info

```{r session_info}
sessionInfo()
```
